---
title: "Predicting Mouse Age (Metagenome)"

---

```{r setup_mlp, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library("dplyr")
library("ggplot2")
cache=T
```

# **Validating** the analysis

A recent Nature Microbiology [study](https://www.nature.com/articles/s41564-025-01963-3) reported a metagenomic examination of mice across the lifespan. They found that the microbiome undergoes an accumulation of stochastic events that are relatively irreversible. This means there is a microbial signal associated with aging, which they leveraged to build machine learning age predictor. They also evaluated the impact of fasting and caloric restriction on the microbiome in these mice.

Their data are available [here](https://github.com/levlitichev/DRiDO_microbiome/tree/main/data).

I was most interested in their Figure 5I, where they tested the hypothesis that caloric restriction/fasting "de-ages" the microbiome. They built a random forest regressor using mice fed *ad libitum* to predict the ages of mice on calorie restriction/fasting regimens. Surprisingly, the predictor guessed their ages were *older* than they actually were, disproving the hypothesis.

Let's see if we can validate these results, and then test other models to see if this was an artefact of the algorithm they chose.

**Note:** I have to mention that this dataset was a dream to work with. The authors clearly made a good effort to make their data as user friendly as possible.

## Loading data

Let's load their genus-level data processed using kraken2.

```{r load_data, echo=F, results=T, plot=T, warning=F, message=F }

mice_data = read.csv("../litichevskiy_validation/kraken_genus_relab_n100x2997.txt", sep="\t")

print(paste0(nrow(mice_data), " genera by ", ncol(mice_data)-1, " samples"))

mice_data[1:5,1:5]


```
It looks like all of the data we need are pre-packed in the feature table. (It's a *lot* of samples!)

## Processing metadata

Let's unpack the metadata from the column names, which appears to follow the format:  Genotype_Diet_Mouse_Time. Of note, we need to subset to these timepoints: 5, 10, 16, 22 and 28 months.

Then we'll check the distributions of samples.

```{r unpack_metadata, echo=F, results=T, plot=T, warning=F, message=F }

# make rownames Genus column, then delete it
rownames(mice_data) = mice_data$genus
mice_data$genus = NULL

# transpose
mice_data = t(mice_data) %>% data.frame()

# extract samples and meta
mice_metadata = data.frame(sample = rownames(mice_data)) %>%
  tidyr::separate(sample, into=c("genotype", "diet", "mouse", "time"), sep="_", remove=F)

# make weeks time
mice_metadata$weeks = gsub("w", "", mice_metadata$time) %>% as.numeric()

# convert weeks to month
mice_metadata$months = mice_metadata$weeks / 4.34

# subset to specified time points: 5, 10, 16, 22 and 28
bound = 0.4 # this bound gives us 2618, as in the paper!
mice_metadata = subset(mice_metadata, 
                       4+bound < months & months < 5+bound | 
                         9+bound < months & months < 10+bound |
                         15+bound < months & months < 16+bound  | 
                         21+bound < months & months < 22+bound |
                         27+bound < months & months < 28+bound )

# subset data
mice_data = mice_data[rownames(mice_data) %in% mice_metadata$sample, ]

# add column to indicate "AD" or "DR"
mice_metadata$class = ifelse(mice_metadata$diet == "AL", "AL", "DR")
table(mice_metadata[,c("genotype", "diet")])

```

Out of the 2618 samples, 528 are "AL", which corresponds to *ad libitum* fed mice, our control group. All of the others refer to some degree of calorie restriction or fasting.

## Data processing

Now let's check the metagenomic data, and see if any additional processing is needed.

```{r check_distributions, echo=F, results=T, plot=T, warning=F, message=F }

# visualize distribution of data
mice_raw_hist = ggplot(reshape2::melt(as.matrix(mice_data)))+
  ggridges::geom_density_ridges(aes(x=value, group=Var2, y=1),
                                  color="black", fill="white", alpha=0.2)+
  theme_minimal()+
  labs(x="Abundance", y="Frequency", title="Metagenomics (Untransformed)")

# calculate pseudocount
mice_pseudocount = min(mice_data[mice_data!=0])/2

# log transform
mice_log = log(mice_data+mice_pseudocount, base=2)

mice_log_hist = ggplot(reshape2::melt(as.matrix(mice_log)))+
  ggridges::geom_density_ridges(aes(x=value, group=Var2, y=1),
                                  color="black", fill="white", alpha=0.2)+
  theme_minimal()+
  labs(x="Log Abundance", y="Frequency", title="Metagenomics (Log-transformed)")

# plot both
mice_raw_hist|mice_log_hist

```

The distributions looks much better after log-transforming with a pseudcount added.

## Building model

Now let's seek to validate their analysis performed in Fig5I. That is, build a random forest regressor using the mice fed *ad libitum* and predict the ages of the calorie restriction/fasting mice. Remember, our goal isn't to achieve high accuracy.

```{r build_models, echo=F, results=T, plot=T, warning=F, message=F }

# first we need to divide the dataset (AL and DR) and keep track of their true ages
mice_al = mice_log[rownames(mice_log) %in% subset(mice_metadata, class == "AL")$sample,]

# to add "age" column, we can just extract the last 4 digits of the rowname!
mice_al$age = gsub("w", "", substr(rownames(mice_al), nchar(rownames(mice_al))-3, nchar(rownames(mice_al)))) %>% as.numeric()

# and repeat for DR
mice_dr = mice_log[rownames(mice_log) %in% subset(mice_metadata, class == "DR")$sample,]
mice_dr$age = gsub("w", "", substr(rownames(mice_dr), nchar(rownames(mice_dr))-3, nchar(rownames(mice_dr)))) %>% as.numeric()

# let's build a predictor
set.seed(25)
mice_rf_model = ranger::ranger(age ~.,
               mice_al)
# extract predictions
mice_al_predictions = mice_rf_model$predictions
# make dataframe
mice_al_pred_df = data.frame(sample = rownames(mice_al),
                             true = mice_al$age,
                             pred = mice_al_predictions,
                             diet = "AL")

# apply predictor and extract predictions
mice_dr_predictions = predict(mice_rf_model, mice_dr)$predictions
mice_dr_pred_df = data.frame(sample = rownames(mice_dr),
                             true = mice_dr$age,
                             pred = mice_dr_predictions,
                             diet = "DR")
# merge
mice_pred_df = rbind(mice_al_pred_df,
                     mice_dr_pred_df) %>% data.frame()

# add months
mice_pred_df = merge(mice_pred_df,
                     mice_metadata[,c("sample", "mouse", "months")], by="sample")
mice_pred_df$month = round(mice_pred_df$months, digits=0)

# calculate median per month
mice_pred_median_df = mice_pred_df %>%
  group_by(month, diet) %>%
  mutate(med.true = median(true/4.34),
         med.pred = median(pred/4.34)) %>%
  dplyr::select(diet, month, med.true, med.pred) %>% distinct() %>% data.frame()

# visualize
mice_pred_df_plot = ggplot(mice_pred_df)+ # convert weeks to months
    # add regression line
    geom_smooth(method="lm", data=subset(mice_pred_df, diet == "DR"),
      aes(x = true/4.34, y = pred/4.34, color="DR"))+
  geom_smooth(method="lm", data=subset(mice_pred_df, diet == "AL"),
      aes(x = true/4.34, y = pred/4.34, color="AL"))+
  # add points
  geom_jitter(data=subset(mice_pred_df, diet == "DR"),
       aes(x = true/4.34, y = pred/4.34, fill="DR"), width=0.6, shape=21, alpha=1)+
  geom_jitter(data=subset(mice_pred_df, diet == "AL"),
       aes(x = true/4.34, y = pred/4.34, fill="AL"), shape=21,alpha=1)+
  # add medians
  geom_segment(data=subset(mice_pred_median_df, diet == "AL") %>% group_by(),
       aes(x = (med.true)-1.5, xend = (med.true)+1.5,
           y = med.pred, yend = med.pred, color="AL"),size=2, alpha=1)+
    geom_segment(data=subset(mice_pred_median_df, diet == "DR"),
       aes(x = (med.true)-1.5, xend = (med.true)+1.5,
           y = med.pred, yend = med.pred, color="DR"),size=2, alpha=1)+
  scale_x_continuous(breaks = c(5, 10, 16, 22, 28))+
    scale_y_continuous(breaks = c(5, 10, 16, 22, 28))+
  theme_minimal()+
  guides(color="none")+
  labs(x="Actual age (months)",
       y="Predicted age (months)",
       fill = "Diet")

mice_pred_df_plot

```

I've added a regression line and lines indicating the median predicted value per diet group.

It looks quite clear that mice from the DR (dietary restriction) group are being predicted to have an older age than what is true.

# **Extending** the analysis

## Part 1: Fixing pseudoreplicates

If we examined the data we're working with closely, we'll have recognized the author committed one of the common mistakes of machine learning: [pseudoreplicates](https://pdobrano25.github.io/ml_website/overfitcheck.html#Pseudoreplicates).

Since they were trying to make a point about a previous hypothesis, rather than build a generalizable model, I wouldn't consider this a grievous case of overfitting. However, I would like to see how a properly engineered random forest would perform. Namely, was their upshift in predictions actually an artefact of overfitting.

So, let's build a stratified random forest such that only 1 sample per mouse is selected for fold. We'll repeat 15 times and take the mean predictions across the stratifications. 

```{r stratify_ranger, echo=F, results=T, plot=T, warning=F, message=F }

# establish stratified folds
# 1, list out mice (in order of AL data)
mouse.subset <- paste("m", mice_metadata[match(rownames(mice_al), mice_metadata$sample),]$mouse, sep="")
# 2, select 1 sample per mouse
mouse.subset.list = do.call(rbind, lapply(1:15, function(iter){
  set.seed(iter)
  data.frame(mouse = mouse.subset) %>%
    mutate(sample = rownames(.)) %>%
    group_by(mouse) %>%
    slice_sample(n=1) %>%
    mutate(iter = iter)
}))
# now loop through iters using stratifications
# take mean of 5 fold predictions

if(cache == F){
# let's build a predictor
mice_pred_stratified = do.call(rbind, lapply(1:15, function(iter){
  mouse.subset.list.subset = subset(mouse.subset.list, iter == iter)
  data.subset = mice_al[as.numeric(mouse.subset.list$sample),]
  set.seed(25)
  mice_rf_model = ranger::ranger(age ~.,
               data.subset)
  # extract predictions
  mice_al_predictions = mice_rf_model$predictions
  # make dataframe
  mice_al_pred_df = data.frame(sample = rownames(data.subset),
                             true = data.subset$age,
                             pred = mice_al_predictions,
                             diet = "AL")

  # apply predictor and extract predictions
  mice_dr_predictions = predict(mice_rf_model, mice_dr)$predictions
  mice_dr_pred_df = data.frame(sample = rownames(mice_dr),
                             true = mice_dr$age,
                             pred = mice_dr_predictions,
                             diet = "DR")
  # merge
  mice_pred_df = rbind(mice_al_pred_df,
                       mice_dr_pred_df) %>% data.frame()
  mice_pred_df$iter = iter
  mice_pred_df
}))

saveRDS(mice_pred_stratified, "../saved/mice_pred_stratified.Rds")

}

mice_pred_stratified = readRDS("../saved/mice_pred_stratified.Rds")


# take mean predictions
mice_pred_stratified_mean = mice_pred_stratified %>%
    mutate(sample = gsub("\\..*", "", sample)) %>%
  group_by(sample) %>%
  mutate(mean.pred = mean(pred)) %>%
  dplyr::select(-pred, -iter) %>% distinct() %>% data.frame()

# add months
mice_pred_stratified_mean = merge(mice_pred_stratified_mean,
                     mice_metadata[,c("sample", "mouse", "months")], by="sample")
mice_pred_stratified_mean$month = round(mice_pred_stratified_mean$months, digits=0)

# calculate median per month
mice_pred_stratified_mean_df = mice_pred_stratified_mean %>%
  group_by(month, diet) %>%
  mutate(med.true = median(true/4.34),
         med.pred = median(mean.pred/4.34)) %>%
  dplyr::select(diet, month, med.true, med.pred) %>% distinct() %>% data.frame()


# visualize
mice_pred_df_plot2 = ggplot(mice_pred_stratified_mean)+ # convert weeks to months
  # add regression line
  geom_smooth(method="lm", data=subset(mice_pred_stratified_mean, diet == "DR"),
       aes(x = true/4.34, y = mean.pred/4.34, color="DR"))+
  geom_smooth(method="lm", data=subset(mice_pred_stratified_mean, diet == "AL"),
       aes(x = true/4.34, y = mean.pred/4.34, color="AL"))+
  # add points
  geom_jitter(data=subset(mice_pred_stratified_mean, diet == "DR"),
       aes(x = true/4.34, y = mean.pred/4.34, fill="DR"), width=0.6, shape=21, alpha=1)+
  geom_jitter(data=subset(mice_pred_stratified_mean, diet == "AL"),
       aes(x = true/4.34, y = mean.pred/4.34, fill="AL"), shape=21,alpha=1)+
  # add medians
  geom_segment(data=subset(mice_pred_stratified_mean_df, diet == "AL") %>% group_by(),
       aes(x = (med.true)-1.5, xend = (med.true)+1.5,
           y = med.pred, yend = med.pred, color="AL"),size=2, alpha=1)+
    geom_segment(data=subset(mice_pred_stratified_mean_df, diet == "DR"),
       aes(x = (med.true)-1.5, xend = (med.true)+1.5,
           y = med.pred, yend = med.pred, color="DR"),size=2, alpha=1)+
  scale_x_continuous(breaks = c(5, 10, 16, 22, 28))+
  scale_y_continuous(breaks = c(5, 10, 16, 22, 28))+
  theme_minimal()+
  guides(color="none")+
  labs(x="Actual age (months)",
       y="Predicted age (months)",
       fill = "Diet",
       title="Pseudoreplicates Removed")

mice_pred_df_plot+labs(title="Pseudoreplicates Kept") | mice_pred_df_plot2
```

Removing pseudoreplicates tells a very different story! Most striking is the change in *slope* of the DR predictions. There is a big upshift in predictions at the first time point (5 months), but it gets neutralized around 16 months, and then reverses at 28 months. Also, it's not a linear shift. Predictions plateau between 16 and 22 months, suggesting the calorie restriction/fasting does not "age" evenly across the lifespan, and may actually *"stop aging" after middle age*.

Another detail to notice, is the AL predictions become more accurate. When pseudoreplicates are kept, 5 month samples are predicted to be ~10 months, and 28 month samples are predicted to be ~19 months. When pseudoreplicates are removed, 5 months = 5 months, and 28 months = 28 months. This suggests the original AL predictions were **downshifted**, likely overfitting the pseudoreplicates towards their mean age value, and removing pseudoreplicates removed this bias.

Our summary statistic of y-intercept is no longer as useful, and consequently, it's unclear how to proceed. We aren't solving an "optimization" problem any more. 

So, let's test other models and just see if the overall conclusions change.

## Part 2. Additional models

We'll build these models

```{r}
ml.models = c("gbm","glmnet","ranger","svmLinear","svmRadial","xgbTree")
```

and consider each of the 15 stratifications as an iteration, and take the mean.

```{r stratify_models_loop, echo=F, results=T, plot=T, warning=F, message=F }

# build helper function to extract predictions
extract_best_pred <- function(model) {
  if (!inherits(model, "train")) stop("Input must be a caret train object")
  best_tune <- model$bestTune
  tune_cols <- names(best_tune)
  best_preds <- model$pred
  for (col in tune_cols) {
    best_preds <- best_preds[best_preds[[col]] == best_tune[[col]], ]
  }
  # take average pred across folds
  best_preds = best_preds %>%
    group_by(rowIndex) %>%
    mutate(mean.pred = mean(pred)) %>%
    dplyr::select(mean.pred, obs, rowIndex) %>% distinct()
  return(best_preds)
}

# loop through models
if(cache == F){

  t1 <- Sys.time()
mice_pred_stratified_loop = do.call(rbind, lapply(ml.models, function(algo){
  # establish stratified folds
  # 1, list out mice (in order of AL data)
  mouse.subset <- paste("m", mice_metadata[match(rownames(mice_al), mice_metadata$sample),]$mouse, sep="")
  # 2, select 1 sample per mouse
  mouse.subset.list = do.call(rbind, lapply(1:15, function(iter){
    set.seed(iter)
    data.frame(mouse = mouse.subset) %>%
      mutate(sample = rownames(.)) %>%
      group_by(mouse) %>%
      slice_sample(n=1) %>%
      mutate(iter = iter)
  }))
# now loop through iters using stratifications
# take mean of 5 fold predictions

# let's build a predictor
mice_pred_stratified = do.call(rbind, lapply(1:15, function(iter){
  
    print(paste0(algo, " iter ", iter))

  
  mouse.subset.list.subset = subset(mouse.subset.list, iter == iter)
  data.subset = mice_al[as.numeric(mouse.subset.list$sample),]
  
  set.seed(25)
  mice_model = caret::train(
    age ~ .,
    data = data.subset,
    trControl = caret::trainControl(method="cv", 
                                    number = 5,
                                    savePred=T),
    metric = "RMSE",
    method = algo,
    preProc = c("center", "scale")) %>% suppressWarnings()
  
  # extract predictions
  mice_al_predictions = extract_best_pred(mice_model)
  # arrange by sample
  mice_al_predictions = mice_al_predictions %>% arrange(rowIndex)
  
  # make dataframe
  mice_al_pred_df = data.frame(sample = rownames(data.subset),
                             true = data.subset$age,
                             pred = mice_al_predictions$mean.pred,
                             diet = "AL")

  # apply predictor and extract predictions
  mice_dr_predictions = predict(mice_model, mice_dr)
  
  mice_dr_pred_df = data.frame(sample = rownames(mice_dr),
                             true = mice_dr$age,
                             pred = mice_dr_predictions,
                             diet = "DR")
  # merge
  mice_pred_df = rbind(mice_al_pred_df,
                       mice_dr_pred_df) %>% data.frame()
  mice_pred_df$iter = iter
  mice_pred_df
}))


# take mean predictions
mice_pred_stratified_mean = mice_pred_stratified %>%
  mutate(sample = gsub("\\..*", "", sample)) %>%
  group_by(sample) %>%
  mutate(mean.pred = mean(pred)) %>%
  dplyr::select(-pred, -iter) %>% distinct() %>% data.frame()

mice_pred_stratified_mean$model = algo

mice_pred_stratified_mean

}))

  t2 <- Sys.time()
  # ~ 2.2 hours
  t2-t1
  
saveRDS(mice_pred_stratified_loop, "../saved/mice_pred_stratified_loop.Rds")

}

mice_pred_stratified_loop = readRDS("../saved/mice_pred_stratified_loop.Rds")

# plot


# add months
mice_pred_stratified_loop = merge(mice_pred_stratified_loop,
                     mice_metadata[,c("sample", "mouse", "months")], by="sample")
mice_pred_stratified_loop$month = round(mice_pred_stratified_loop$months, digits=0)

# calculate median per month
mice_pred_stratified_loop_mean = mice_pred_stratified_loop %>%
  group_by(month, diet, model) %>%
  mutate(med.true = median(true/4.34),
         med.pred = median(mean.pred/4.34)) %>%
  dplyr::select(model, diet, month, med.true, med.pred) %>% distinct() %>% data.frame()


# visualize
mice_pred_loop_plot = ggplot(mice_pred_stratified_loop)+ # convert weeks to months
  # add regression lines
    geom_smooth(method="lm", data=subset(mice_pred_stratified_loop, diet == "DR"),
       aes(x = true/4.34, y = mean.pred/4.34, color="DR"))+
  geom_smooth(method="lm", data=subset(mice_pred_stratified_loop, diet == "AL"),
       aes(x = true/4.34, y = mean.pred/4.34, color="AL"))+
  # add points
  geom_jitter(data=subset(mice_pred_stratified_loop, diet == "DR"),
       aes(x = true/4.34, y = mean.pred/4.34, fill="DR"), width=0.6, shape=21, alpha=1)+
  geom_jitter(data=subset(mice_pred_stratified_loop, diet == "AL"),
       aes(x = true/4.34, y = mean.pred/4.34, fill="AL"), shape=21,alpha=1)+
  # add medians
  geom_segment(data=subset(mice_pred_stratified_loop_mean, diet == "AL") %>% group_by(),
       aes(x = (med.true)-1.5, xend = (med.true)+1.5,
           y = med.pred, yend = med.pred, color="AL"),size=2, alpha=1)+
    geom_segment(data=subset(mice_pred_stratified_loop_mean, diet == "DR"),
       aes(x = (med.true)-1.5, xend = (med.true)+1.5,
           y = med.pred, yend = med.pred, color="DR"),size=2, alpha=1)+
  scale_x_continuous(breaks = c(5, 10, 16, 22, 28))+
  scale_y_continuous(breaks = c(5, 10, 16, 22, 28))+
  theme_minimal()+
  guides(color="none")+
  labs(x="Actual age (months)",
       y="Predicted age (months)",
       fill = "Diet",
       title="Pseudoreplicates Removed")+
  facet_wrap(~model)

mice_pred_loop_plot

```

There is some model-specificity with this effect. GBM, random forest, svmRadial, and XGBoost predictions plateau at 22 months, while elastic net and svmLinear plateau 6 months later at the final time point. Interestingly, these models can be defined as "non-linear" vs "linear" models according to when the predictions plateau.

Nonetheless, these results leave a question mark after the conclusions made in the original study.

# Conclusions

To conclude, we found that one of the study's conclusions changes when machine learning is engineered properly. 

