---
title: "Discriminating IBD using Biopsy Transcriptomes"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache=TRUE)
library("dplyr")
library("ggplot2")

```


[Argmann et al](https://gut.bmj.com/content/72/7/1271.long) performed RNAseq on a mucosal biopsies and blood from a large cohort of IBD and non-IBD controls. Their primary goal was to construct a "molecular inflammation" score that correlates with established measurements of inflammation. For reasons outlined below, I was unable to validate this analysis. 

Their publicly available [dataset](https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE193677) enables testing other hypotheses using machine learning.

Below, we process the dataset and apply machine learning to solve two tasks.

- **Predict IBD** in inflamed and non-inflamed rectum biopsy samples

- **Predict IBD in non-inflamed** rectum and ileum biopsy samples using an **ensemble model**.

# **Re-analyzing** the data

# Loading data

First, let's check the total number of samples and features:

```{r, data_loading, echo=F}

# SLOW; cannot be sped up

# load normalized count data (from GEO: https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE193677)
# click adj_counts
sinai_data = read.csv("../sinai_validation/GSE193677_norm_counts_TPM_GRCh38.p13_NCBI.tsv", sep="\t")

print(paste(dim(sinai_data)[1], " samples by ", dim(sinai_data)[2], " features", sep=""))

```

Next, we will assign gene names to transcripts. Not all transcripts can be annotated, so we will remove them. This is not necessarily the optimal decision for the purpose of building the most accurate classifier, but it helps render interpretability.

```{r, keep_annotated, echo=F}
# create gene annotations: https://www.gencodegenes.org/human/release_46.html
annotation1 = read.csv("../sinai_validation/gencode.v46.metadata.EntrezGene", sep="", header=F)
annotation2 = read.csv("../sinai_validation/gencode.v46.metadata.HGNC", sep="", header=F)
# merge
annotations = merge(annotation1, annotation2, by="V1")
# keep needed columns
annotations = annotations[,c("V2.x", "V2.y")]
colnames(annotations) = c("feature", "name")

# swap in data
sinai_data$GeneID = annotations[match(sinai_data$GeneID, annotations$feature),]$name
# remove NA genes
sinai_data = sinai_data[!is.na(sinai_data$GeneID),]
rownames(sinai_data) = sinai_data$GeneID
sinai_data$GeneID = NULL

print(paste0(ncol(sinai_data), " features remain", sep=""))


```


## Prevalence filtering 

Next, we'll remove features that are present in 3 samples or fewer.

```{r, prevalence_filter, echo=F}

# convert non0 value to 1
sinai_data_prev = sinai_data
sinai_data_prev[sinai_data_prev != 0] = 1
# remove features with 3 or fewer 1's
sinai_data = sinai_data[rowSums(sinai_data_prev) > 3,]

# let's also transpose the data
sinai_data = t(sinai_data) %>% data.frame()

print(paste("Removed: ", sum(rowSums(sinai_data_prev) <= 3), " features"))


```

This removed a relatively small number of features.

## Distribution check

Lastly, we'll check the distributions of the data to ensure they are approximately normally distributed prior to visualizing with PCA. Let's randomly select 50 genes to get a sense of the distributions:

```{r, data_distributions, echo=F, results = T, message=F, warning=F}

# plot distributions of 50 random genes
set.seed(25)
sample(sinai_data, size=50, replace=F)%>% as.matrix() %>%
  reshape2::melt() %>%
  ggplot(aes(x=value, y = 1))+
  ggridges::geom_density_ridges(aes(group = Var2), color="black", fill="white", alpha=0.2)+
  theme_minimal()+theme(legend.position="none")+
  labs(x="Adjusted Count", y="Frequency")

```

These data appear severely [right-skewed](https://en.wikipedia.org/wiki/Skewness). 

## Log-transformation

Let's apply a log-transform (with pseudocount).

```{r, data_distributions_post_log, echo=F, results = F, plot=T, message=F, warning=F}

sinai_pseudo = min(sinai_data[sinai_data!=0])/2

sinai_data_log = log(sinai_data + sinai_pseudo, base=2)
# plot distributions of 50 random genes
set.seed(25)
sample(sinai_data_log, size=50, replace=F)%>% as.matrix() %>%
  reshape2::melt() %>%
  ggplot(aes(x=value, y = 1))+
  ggridges::geom_density_ridges(aes(group = Var2), color="black", fill="white", alpha=0.2)+
  theme_minimal()+theme(legend.position="none")+
  labs(x="Log-transformed Adjusted Count", y="Frequency")

```

These look significantly more normally-distributed than before.

We have finished pre-processing our data. Here are what the first 4 rows and columns look like:


```{r, sense_of_data, echo=F}
sinai_data_log[1:4, 1:4]
```


```{r, mapping, echo=F}

# load mapping data (from: https://www.ncbi.nlm.nih.gov/sra?term=SRP354989)
# click "Send to: txt"
sinai_meta = read.csv("../sinai_validation/sra_result.csv", sep=",")
# process metadata:
sinai_meta$Experiment.Accession = NULL
sinai_meta = sinai_meta %>%
  tidyr::separate(col=`Experiment.Title`, into=c("Sample", "other"), sep=": ", remove=T) %>%
    tidyr::separate(col=other, into=c("Patient", "other"), sep="_Biopsy_", remove=T) %>%
      tidyr::separate(col=other, into=c("other", "Info"), sep=", ", remove=T) %>%
        tidyr::separate(col=Info, into=c("Disease", "other"), sep=",", remove=T) %>%
        mutate(other = gsub("\\;.*", "", other)) %>%
        tidyr::separate(col=other, into=c("Location", "Status", "tissue"), sep=" ", remove=T) %>%
          mutate(Disease = gsub(" participants", "", Disease)) %>%
          mutate(Patient = gsub("MSCCR_reGRID_", "P", Patient)) %>%
          dplyr::select(Sample, Patient, Disease, Location, Status) %>% distinct()

sinai_meta %>%
  #subset(Location == "Rectum" & Status == "NonI") %>% 
  select(Patient, Disease) %>%
  distinct() %>% 
  select(Disease) %>% table() %>%
  data.frame() %>%
  `colnames<-`(c("disease", "sample count"))

```

## Principal Component Analysis 

Now let's visualize how the samples relate to each other after dimension reduction (using PCA): 

```{r, all_data_pca, echo = F, results = T, warning = F}

cached = T

if(cached == F){
  
# construct PCA
sinai_pca <- prcomp((sinai_data_log), center = TRUE, scale = TRUE)

# extract results
sinai_pca_df <- as.data.frame(sinai_pca$x)  # PCA scores
sinai_pca_df$Sample <- rownames(sinai_pca$x)  # Add sample names for visualization
# add meta
sinai_pca_df = merge(sinai_pca_df,
                   sinai_meta, by= "Sample")

# Variance explained by each principal component
var_exp <- round((sinai_pca$sdev^2 / sum(sinai_pca$sdev^2)) * 100, 2)

# let's reorder the location by biogeography
sinai_pca_df$Location = factor(sinai_pca_df$Location,
                                    levels = c("Ileum", "Cecum", "LeftColon", "Transverse", "RightColon","Sigmoid","Rectum"))

saveRDS(sinai_pca_df, "../saved/sinai_pca_df.Rds")
saveRDS(var_exp, "../saved/var_exp_sinai.Rds")

}

sinai_pca_df <- readRDS("../saved/sinai_pca_df.Rds")
var_exp <- readRDS("../saved/var_exp_sinai.Rds")

ggplot(sinai_pca_df,
       aes(x=PC1, y=PC2))+
  geom_point(data=subset(sinai_pca_df, Status == "I"), shape=21, color="black",
             aes(fill = as.factor("Inflamed"), shape = Status))+
    geom_point(data=subset(sinai_pca_df, Status == "NonI"), shape=23, color="black",
             aes(fill = as.factor("Non-inflamed"), shape = Status))+
  theme_bw()+
  facet_wrap(~Location)+
  labs(x=paste("PC1: ", round(var_exp[1], digits=2), "%", sep=""), 
       y=paste("PC2: ", round(var_exp[2], digits=2), "%", sep=""),
       fill = "Status")

```

The separation by location doesn't look great. Nonetheless, we're going to subset the data to one compartment to simplify the problem. Which compartment has the greatest number of samples with good representation among conditions?

```{r, compare_compartments, echo = F, results = T, plot = F, warning=F}

table(sinai_meta[,c("Location", "Status", "Disease")]) %>% data.frame() %>%
  group_by(Location) %>%
  mutate(sum.location = sum(Freq)) %>%
  group_by(Location, Disease) %>%
    mutate(sum.location.disease = sum(Freq)) %>%
  group_by(Location, Disease, Status) %>%
    mutate(sum.location.disease.status = sum(Freq)) %>% data.frame() %>%
  arrange(-sum.location) %>%
  head() %>%
  `colnames<-`(c("Location", "Status", "Disease", "Samples",
                 "Samples per Location",
                 "Samples per Location per Disease",
                 "Samples Per Location per Disease per Status"))


```

We can see that the rectum compartment has the most samples overall, with high representation per disease and status.

This makes biological sense. The rectum may be inflamed in CD or UC, but the ileum is rarely affected in UC.

## Subset to rectum

Let's subset the data to the rectum and apply PCA again, and take only one sample per patient.

```{r, subset_to_rectum, echo = F, results = F, plot = T, warning=F}

  
# subset meta to Rectum
sinai_rectum_meta = subset(sinai_meta, Location == "Rectum")
# check pseudoreplicates
sinai_rectum_meta$Patient %>% table() %>% range()
# remove pseudoreplicates
sinai_rectum_meta = sinai_rectum_meta %>%
  group_by(Patient) %>%
  slice_head(n = 1)
# check pseudoreplicates
sinai_rectum_meta$Patient %>% table() %>% range()

# subset data to Rectum
sinai_rectum = sinai_data_log[rownames(sinai_data_log) %in% sinai_rectum_meta$Sample,]

cached = T

if(cached ==F){

# apply prevalence filter again
sinai_rectum_prev = (2^sinai_rectum) - sinai_pseudo
# need to round down because of floating integer
sinai_rectum_prev[sinai_rectum_prev == min(sinai_rectum_prev)] = 0
# return to prev filter operation
sinai_rectum_prev[sinai_rectum_prev != 0] = 1
# remove features with 3 or fewer 1's
sinai_rectum = sinai_rectum[,colSums(sinai_rectum_prev) > 3]
dim(sinai_rectum)

# construct PCA
sinai_rectum_pca <- prcomp((sinai_rectum), center = TRUE, scale = TRUE)

# extract results
sinai_rectum_pca_df <- as.data.frame(sinai_rectum_pca$x)  # PCA scores
sinai_rectum_pca_df$Sample <- rownames(sinai_rectum_pca$x)  # Add sample names for visualization
# add meta
sinai_rectum_pca_df = merge(sinai_rectum_pca_df,
                   sinai_rectum_meta, by= "Sample")

# Variance explained by each principal component
var_exp_rectum <- round((sinai_rectum_pca$sdev^2 / sum(sinai_rectum_pca$sdev^2)) * 100, 2)

saveRDS(sinai_rectum_pca_df, "../saved/sinai_rectum_pca_df.Rds")
saveRDS(var_exp_rectum, "../saved/var_exp_rectum.Rds")

}

sinai_rectum_pca_df <- readRDS("../saved/sinai_rectum_pca_df.Rds")
var_exp_rectum <- readRDS("../saved/var_exp_rectum.Rds")

library("patchwork")
ggplot(sinai_rectum_pca_df,
       aes(x=PC1, y=PC2))+
  geom_point(data=subset(sinai_rectum_pca_df, Status == "I"), shape=21, color="black",
             aes(fill = as.factor(1), shape = Status))+
    geom_point(data=subset(sinai_rectum_pca_df, Status == "NonI"), shape=23, color="black",
             aes(fill = as.factor(2), shape = Status))+
  theme_minimal()+theme(legend.position="none")+
  labs(x=paste("PC1: ", round(var_exp_rectum[1], digits=2), "%", sep=""), 
       y=paste("PC2: ", round(var_exp_rectum[2], digits=2), "%", sep=""))+
  # add distributions along axes
  ggplot(sinai_rectum_pca_df, 
         aes(x=PC2, y=1))+
  geom_violin(aes(x=Status, y=PC2, fill=Status), color="black",
              draw_quantiles = 0.5)+
  theme_minimal()+theme(legend.position="none")+
  labs(x="Status", y="PC2")+  
  ggplot(sinai_rectum_pca_df, 
         aes(x=PC1, y=1))+
  coord_flip()+
  geom_violin(aes(x=Status, y=PC1, fill=Status), color="black",
              draw_quantiles = 0.5)+
  theme_minimal()+theme(legend.position="none")+
  labs(x="Status", y="PC1")+ plot_spacer()+
  patchwork::plot_layout(widths=c(4,1), heights=c(3,1))

```

PCA of rectum biopsies shows good separation based on whether the tissue was inflamed or not.

Now we can start asking questions about whether the transcriptome can diagnose inflammation or discriminate disease.

I personally think predicting inflammation is not a useful application of machine learning (bordering on tautological), because a clinician will be able to determine whether a tissue is (endoscopically) inflamed while sampling it. Nonetheless, it poses a straightforward task for machine learning.

## Molecular Inflammation Score

Before launching into machine learning, I wanted to comment on what the authors advanced in their study.

They used linear mixed effects regression to identify genes *upregulated* in inflamed samples, followed by *GSVA* to calculate a single score per sample, called the *Molecular Inflammation Score*.

In their model, they controlled for batch, RNA integrity number, rRNA rate, exonic rate, age, gender, and genetic principal components. However, these are data that are not readily available from GEO.

So, we will move on to applying machine learning instead.

# **Predicting IBD**

For the sake of running this analysis locally, we're going to drastically reduce the number of features used as predictive variables.

## Variance filter

**Feature selection** is an important step in pre-processing data prior to training a model. We have a decently-sized dataset, but I prefer not spending samples performing supervised feature selection (e.g. running t-tests to extract features we suspect to be associated with the outcome). Instead, we'll apply unsupervised feature selection by removing low variance and highly correlated features.

First, let's use caret's build in "nearZeroVar" function. From the [documentation](https://www.rdocumentation.org/packages/caret/versions/6.0-92/topics/nearZeroVar), this function removes features with:
*one unique value (i.e. are zero variance predictors) or predictors that are have both of the following characteristics: they have very few unique values relative to the number of samples and the ratio of the frequency of the most common value to the frequency of the second most common value is large.*

```{r, variance_filter, echo=F}

# first, initialize new dataframe
sinai_rectum_ibd = sinai_rectum %>% data.frame()

# identify nearzerovar
nearzervar_features = caret::nearZeroVar(sinai_rectum_ibd, freqCut = 10/1, uniqueCut = 20)
# length(nearzervar_features)

# remove nearzerovar
sinai_rectum_ibd_filt = sinai_rectum_ibd[,-nearzervar_features]

print(paste("Removed: ", length(nearzervar_features), " features (", round((1 - ncol(sinai_rectum_ibd_filt) / ncol(sinai_rectum_ibd)) *100, digits=2), "%)", sep=""))

```

## Correlated features filter

We can also apply the "findCorrelation" function to eliminate features that are highly correlated (i.e. carry redundant information), and retain a single feature among them to represent the rest. Here, we'll apply an aggressive threshold (Pearson correlation > 0.70).

```{r, correlation_filter, echo=F}

cached = T

if(cached == F){
  
# calculate correlations
cor_matrix <- cor(sinai_rectum_ibd_filt)

# identify highly correlated features
high_cor <- caret::findCorrelation(cor_matrix, cutoff = 0.7)
#length(high_cor)
saveRDS(high_cor, "../saved/high_cor.Rds")

}

high_cor = readRDS("../saved/high_cor.Rds")
# remove correlated features

sinai_rectum_ibd_filt_cor = sinai_rectum_ibd_filt[,-high_cor]

print(paste("Removed: ", length(high_cor), " features (", round((1 - ncol(sinai_rectum_ibd_filt_cor) / ncol(sinai_rectum_ibd_filt)) *100, digits=2), "%)", sep=""))

```

Correlation filtering removed a big chunk of the features.

## Train model

Now, we'll use all rectum biopsies to predict Control from IBD (CD or UC)

We will first use glmnet to establish a proof-of-principle that this prediction can be made. Because we have a good sample size, we'll rebalance the dataset during CV by randomly downsampling to the smallest class.

```{r, build_model, echo=F, warning=F, message=F}

sinai_rectum_ibd_filt_ibd = sinai_rectum_ibd_filt_cor

# add outcome
sinai_rectum_ibd_filt_ibd$Sample = rownames(sinai_rectum_ibd_filt_ibd)
sinai_rectum_ibd_filt_ibd = merge(sinai_rectum_ibd_filt_ibd,
                              sinai_rectum_meta[,c("Sample", "Disease")],
                              by="Sample")
rownames(sinai_rectum_ibd_filt_ibd) = sinai_rectum_ibd_filt_ibd$Sample
sinai_rectum_ibd_filt_ibd$Sample = NULL

# make UC and CD into IBD
sinai_rectum_ibd_filt_ibd$Disease = ifelse(sinai_rectum_ibd_filt_ibd$Disease == "Control", "Control", "IBD")

# check balance
table(sinai_rectum_ibd_filt_ibd$Disease)
# 221 to 660

cached = T

if(cached == F){
  
# build 5x CV model
sinai_ibd_model_train <- caret::train(
  Disease ~ .,
  data = sinai_rectum_ibd_filt_ibd,
  trControl = caret::trainControl(method="cv", 
                                  number = 5,
                                  savePredictions = TRUE,
                                  classProbs = TRUE,
                                  summaryFunction = caret::twoClassSummary,
                                  sampling = "down"),
  metric = "ROC",
  method = "glmnet",
  preProc = c("center", "scale")
)

# let's plot the ROC curve

# select the final model
sinai_ibd_model_train_best = subset(sinai_ibd_model_train$pred, alpha == sinai_ibd_model_train$bestTune$alpha & lambda == sinai_ibd_model_train$bestTune$lambda)

# build ROC curve
roc_curve_all <- pROC::roc(sinai_ibd_model_train_best$obs, sinai_ibd_model_train_best$IBD, levels = levels(sinai_ibd_model_train_best$obs))
# convert ROC Curve to Data Frame
roc_df <- data.frame(
  specificity = (roc_curve_all$specificities),
  sensitivity = roc_curve_all$sensitivities
) %>% arrange((sensitivity))

# calculate 95% confidence interval
roc_curve_ci <- pROC::ci.se(roc_curve_all, specificities = seq(0, 1, l = 100), boot.n = 2000)
roc_curve_ci = data.frame(roc_curve_ci)
colnames(roc_curve_ci) = c("lower", "mean", "upper")
roc_curve_ci$spec = seq(0, 1, l = 100)

saveRDS(sinai_ibd_model_train_best, "../saved/sinai_ibd_model_train_best.Rds")
saveRDS(roc_df, "../saved/roc_df.Rds")
saveRDS(roc_curve_ci, "../saved/roc_curve_ci.Rds")

}

sinai_ibd_model_train_best = readRDS("../saved/sinai_ibd_model_train_best.Rds")
roc_df = readRDS("../saved/roc_df.Rds")
roc_curve_ci = readRDS("../saved/roc_curve_ci.Rds")

# recalculate AUC
roc_curve_all <- pROC::roc(sinai_ibd_model_train_best$obs, sinai_ibd_model_train_best$IBD, levels = levels(sinai_ibd_model_train_best$obs))

# plot Using ggplot2
ggplot() +  
  # 95% Confidence Interval as a shaded region
  geom_ribbon(data = roc_curve_ci, aes(x = rev(spec), ymin = lower, ymax = upper), 
              fill = "blue", alpha = 0.2) +
  # ROC Curve
  geom_line(data = roc_df, aes(x = 1-specificity, y = sensitivity), color = "black", size = 1) +

  # Add diagonal reference line (random classifier)
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray") +
  # Labels and theme
  labs(title = paste("ROC Curve with 95% CI", " (AUC: ", round(roc_curve_all$auc, digits=3), ")", sep=""), x = "1 - Specificity", y = "Sensitivity") +
  theme_bw()
```

Even after aggressive feature selection and downsampling, elastic net can predict IBD from controls with an AUC of 0.906. Quite an impressive looking ROC, albeit for internal validation.

## Questioning the utility

I stated before that I don't consider this a very useful problem to solve. The clinician can make a determination of IBD or not well enough. In fact, I suspect that the high AUC is leveraging the fact that inflammation (visible on endoscopy) is represented by a highly discriminative transcriptomic signature alone, and this model would fail in discriminating IBD in non-inflamed samples. Let's explore this suspicion.

```{r, model_isnt_useful, echo=F, results=T, warning=F, message=F}

# extract mean predictions and link to inflammation status
sinai_ibd_model_mean = sinai_ibd_model_train_best %>%
  group_by(rowIndex) %>%
  mutate(mean.prob = mean(IBD)) %>%
  dplyr::select(rowIndex, mean.prob) %>% distinct() %>% data.frame() %>%
  arrange(rowIndex) %>%
  mutate(Sample = rownames(sinai_rectum_ibd_filt_ibd))

sinai_ibd_model_mean = merge(sinai_ibd_model_mean,
                             sinai_rectum_meta, by="Sample")

sinai_ibd_model_mean$Disease = factor(sinai_ibd_model_mean$Disease, levels=c("Control", "CD", "UC"))
sinai_ibd_model_mean$Status = factor(ifelse(sinai_ibd_model_mean$Status == "NonI", "Non-inflamed", "Inflamed"), levels=c("Non-inflamed", "Inflamed"))
sinai_ibd_model_mean$Status_color = factor(sinai_ibd_model_mean$Status, levels=c("Inflamed", "Non-inflamed"))

ggplot(sinai_ibd_model_mean,
       aes(x=Disease, y=mean.prob))+
  geom_violin(draw_quantiles= 0.5) +
  ggbeeswarm::geom_beeswarm(shape=21, aes(fill = Status_color), color="black")+
  geom_hline(yintercept=0.5, linetype=2, alpha=0.5)+
  theme_minimal()+theme(legend.position="none")+
  facet_wrap(~Status, scales="free_x")+
  labs(x="Disease", y="Mean Probability of IBD")

```

Here, it is evident that the model is most discriminative if the tissue is inflamed. There does appear to be a signal in the non-inflamed UC rectum, but less so for CD.

# **Predicting IBD with non-inflamed transcriptomes**

We also have ileum data, which is more specific for CD. Since the rectum should be more specific to UC, a hypothesis can be raised that I have not seen tested before:

Could an **ensemble** of 2 machine learning models (one trained on the rectum and one trained on the ileum) diagnose IBD from controls *in the absence of inflammation* ?

```{r, preparing_ensemble_data, echo=F, results=F, plot=T, warning=F, message=F}

# first, we need to subset the RECTUM and ILEUM data to non-inflamed samples

sinai_joint_meta = subset(sinai_meta, Location %in% c("Rectum", "Ileum") & 
                            Status == "NonI") %>%
  group_by(Patient, Location) %>%
  slice_head(n = 1)
# table(sinai_joint_meta[,c("Disease","Location")])

# we're going to use matched samples, which will shrink our dataset considerably
matching_samples = sinai_joint_meta[,c("Patient")] %>% table() %>% data.frame() %>% subset(Freq>1)
# apply filter
sinai_joint_meta = subset(sinai_joint_meta, Patient %in% matching_samples$Patient)
# table(sinai_joint_meta[,c("Disease", "Location")])

# now reformat feature tables to be structured identically

sinai_joint_r = sinai_data_log[rownames(sinai_data_log) %in% subset(sinai_joint_meta, Location == "Rectum")$Sample,]
# dim(sinai_joint_r)

sinai_joint_i = sinai_data_log[rownames(sinai_data_log) %in% subset(sinai_joint_meta, Location == "Ileum")$Sample,]
# dim(sinai_joint_i)

# replace rownames with patient number
rownames(sinai_joint_r) = sinai_joint_meta[match(rownames(sinai_joint_r), sinai_joint_meta$Sample),]$Patient

rownames(sinai_joint_i) = sinai_joint_meta[match(rownames(sinai_joint_i), sinai_joint_meta$Sample),]$Patient

cache = T

if(cache == F){
# filter features per dataset
# rectum
nearzervar_features_r = caret::nearZeroVar(sinai_joint_r, freqCut = 10/1, uniqueCut = 20)
sinai_joint_r_filt = sinai_joint_r[,-nearzervar_features_r]
cor_matrix_r <- cor(sinai_joint_r_filt)
high_cor_r <- caret::findCorrelation(cor_matrix_r, cutoff = 0.7)
sinai_joint_r_filt_cor = sinai_joint_r_filt[,-high_cor_r]
# ileum
nearzervar_features_i = caret::nearZeroVar(sinai_joint_i, freqCut = 10/1, uniqueCut = 20)
sinai_joint_i_filt = sinai_joint_i[,-nearzervar_features_i]
cor_matrix_i <- cor(sinai_joint_i_filt)
high_cor_i <- caret::findCorrelation(cor_matrix_i, cutoff = 0.7)
sinai_joint_i_filt_cor = sinai_joint_i_filt[,-high_cor_i]

saveRDS(sinai_joint_r_filt_cor, "../saved/sinai_joint_r_filt_cor.Rds")
saveRDS(sinai_joint_i_filt_cor, "../saved/sinai_joint_i_filt_cor.Rds")

}

sinai_joint_r_filt_cor = readRDS("../saved/sinai_joint_r_filt_cor.Rds")
sinai_joint_i_filt_cor = readRDS("../saved/sinai_joint_i_filt_cor.Rds")

print(paste0("Rectum features: ", ncol(sinai_joint_r_filt_cor), ". Ileum features: ", ncol(sinai_joint_i_filt_cor), "."))

```

To review, we've limited the dataset patients with matching ileum and rectum samples. We've removed [pseudoreplicates](https://pdobrano25.github.io/ml_website/overfitcheck.html#Pseudoreplicates), such that each patient is only represented once. Lastly, we've filtered out low variance and highly correlated (R > 0.70) features.

Now we can continue on to building the ensemble. We'll build separate classifiers: one for predicting IBD in ileum samples, and one for predicting IBD in rectum samples.

Of note: we will need a consistent CV scheme that both models will use. Because we are also downsampling, we will pull out the CV scheme from the first model and apply it to the second (and ensemble).

```{r, ensemble_rectum, echo=F, results=F, plot=T, warning=F, message=F}

cached = T

if(cached == F){
  
# initialize data
sinai_r = sinai_joint_r_filt_cor

# add outcome
sinai_r$Patient = rownames(sinai_r)
sinai_r = merge(sinai_r,
                subset(sinai_joint_meta, Location == "Rectum")[,c("Patient", "Disease")],
                by="Patient")
rownames(sinai_r) = sinai_r$Pateint
sinai_r$Patient = NULL

# make UC and CD into IBD
sinai_r$Disease = ifelse(sinai_r$Disease == "Control", "Control", "IBD") %>% as.factor()

# check balance
table(sinai_r$Disease)
# 111 to 150

# check dim
dim(sinai_r)

# build 5x CV model
set.seed(25)
sinai_r_model_train <- caret::train(
  Disease ~ .,
  data = sinai_r,
  trControl = caret::trainControl(method="cv", 
                                  number = 5,
                                  savePredictions = TRUE,
                                  classProbs = TRUE,
                                  summaryFunction = caret::twoClassSummary,
                                  sampling = "down"),
  metric = "ROC",
  method = "glmnet",
  preProc = c("center", "scale")
)

# save CV scheme for next model
ensemble_cv_scheme = sinai_r_model_train$control$index

# select the final model
sinai_r_model_train_best = subset(sinai_r_model_train$pred, alpha == sinai_r_model_train$bestTune$alpha & lambda == sinai_r_model_train$bestTune$lambda)

# build ROC curve
roc_curve_all_r <- pROC::roc(sinai_r_model_train_best$obs, sinai_r_model_train_best$IBD, levels = levels(sinai_r_model_train_best$obs))
# convert ROC Curve to Data Frame
roc_df_r <- data.frame(
  specificity = (roc_curve_all_r$specificities),
  sensitivity = roc_curve_all_r$sensitivities
) %>% arrange((sensitivity))

# calculate 95% confidence interval
roc_curve_ci_r <- pROC::ci.se(roc_curve_all_r, specificities = seq(0, 1, l = 100), boot.n = 2000)
roc_curve_ci_r = data.frame(roc_curve_ci_r)
colnames(roc_curve_ci_r) = c("lower", "mean", "upper")
roc_curve_ci_r$spec = seq(0, 1, l = 100)

saveRDS(sinai_r_model_train_best, "../saved/sinai_r_model_train_best.Rds")
saveRDS(roc_df_r, "../saved/roc_df_r.Rds")
saveRDS(roc_curve_ci_r, "../saved/roc_curve_ci_r.Rds")

}

sinai_r_model_train_best = readRDS("../saved/sinai_r_model_train_best.Rds")
roc_df_r = readRDS("../saved/roc_df_r.Rds")
roc_curve_ci_r = readRDS("../saved/roc_curve_ci_r.Rds")

# recalculate AUC (because I didn't save it the first time!)
roc_curve_all_r <- pROC::roc(sinai_r_model_train_best$obs, sinai_r_model_train_best$IBD, levels = levels(sinai_r_model_train_best$obs))


# plot Using ggplot2
ggplot() +  
  # 95% Confidence Interval as a shaded region
  geom_ribbon(data = roc_curve_ci_r, aes(x = rev(spec), ymin = lower, ymax = upper), 
              fill = "blue", alpha = 0.2) +
  # ROC Curve
  geom_line(data = roc_df_r, aes(x = 1-specificity, y = sensitivity), color = "black", size = 1) +

  # Add diagonal reference line (random classifier)
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray") +
  # Labels and theme
  labs(title = paste("Rectum (Non-inflamed)\nROC Curve with 95% CI", " (AUC: ", round(roc_curve_all_r$auc, digits=3), ")", sep=""), x = "1 - Specificity", y = "Sensitivity") +
  theme_bw()

```
```{r, is_r_model_useful, echo=F, results=F, plot=T, warning=F, message=F}

# extract mean predictions and link to inflammation status
sinai_r_model_mean = sinai_r_model_train_best %>%
  group_by(rowIndex) %>%
  mutate(mean.prob = mean(IBD)) %>%
  dplyr::select(rowIndex, mean.prob) %>% distinct() %>% data.frame() %>%
  arrange(rowIndex) %>%
  mutate(Patient = rownames(sinai_joint_r))

sinai_r_model_mean = merge(sinai_r_model_mean,
                             subset(sinai_joint_meta, Location == "Rectum"), by="Patient")

sinai_r_model_mean$Disease = factor(sinai_r_model_mean$Disease, levels=c("Control", "CD", "UC"))
sinai_r_model_mean$Status = factor(sinai_r_model_mean$Status, levels=c("NonI", "I"))

sinai_r_model_mean$Status = factor(ifelse(sinai_r_model_mean$Status == "NonI", "Non-inflamed", "Inflamed"), levels=c("Non-inflamed", "Inflamed"))

ggplot(sinai_r_model_mean,
       aes(x=Disease, y=mean.prob))+
  geom_violin(draw_quantiles= 0.5) +
  ggbeeswarm::geom_beeswarm(shape=21, aes(fill = Disease), color="black")+
  geom_hline(yintercept=0.5, linetype=2, alpha=0.5)+
  theme_minimal()+theme(legend.position="none")+
  facet_wrap(~Status, scales="free_x")+
  labs(x="Disease", y="Mean Probability of IBD", title="Non-Inflamed Rectum Model")


```

This rectum-only model is doing what we want it to: predict CD or UC from non-inflamed rectal tissues. 

Since inflammation in rectal tissue is not specific to CD or UC, we'll build a similar classifier using ileum samples.

```{r, ensemble_ileum, echo=F, results=F, plot=T, warning=F, message=F}

cached = T

if(cached == F){
  
# initialize data
sinai_i = sinai_joint_i_filt_cor

# add outcome
sinai_i$Patient = rownames(sinai_i)
sinai_i = merge(sinai_i,
                subset(sinai_joint_meta, Location == "Ileum")[,c("Patient", "Disease")],
                by="Patient")
rownames(sinai_i) = sinai_i$Pateint
sinai_i$Patient = NULL

# make UC and CD into IBD
sinai_i$Disease = ifelse(sinai_i$Disease == "Control", "Control", "IBD") %>% as.factor()

# check balance
table(sinai_i$Disease)
# 111 to 150

# check dim
dim(sinai_i)

# build 5x CV model
set.seed(25)
sinai_i_model_train <- caret::train(
  Disease ~ .,
  data = sinai_i,
  trControl = caret::trainControl(method="cv", 
                                  index = ensemble_cv_scheme,
                                  savePredictions = TRUE,
                                  classProbs = TRUE,
                                  summaryFunction = caret::twoClassSummary,
                                  sampling = "down"),
  metric = "ROC",
  method = "glmnet",
  preProc = c("center", "scale")
)

# select the final model
sinai_i_model_train_best = subset(sinai_i_model_train$pred, alpha == sinai_i_model_train$bestTune$alpha & lambda == sinai_i_model_train$bestTune$lambda)

# build ROC curve
roc_curve_all_i <- pROC::roc(sinai_i_model_train_best$obs, sinai_i_model_train_best$IBD, levels = levels(sinai_i_model_train_best$obs))
# convert ROC Curve to Data Frame
roc_df_i <- data.frame(
  specificity = (roc_curve_all_i$specificities),
  sensitivity = roc_curve_all_i$sensitivities
) %>% arrange((sensitivity))

# calculate 95% confidence interval
roc_curve_ci_i <- pROC::ci.se(roc_curve_all_i, specificities = seq(0, 1, l = 100), boot.n = 2000)
roc_curve_ci_i = data.frame(roc_curve_ci_i)
colnames(roc_curve_ci_i) = c("lower", "mean", "upper")
roc_curve_ci_i$spec = seq(0, 1, l = 100)


saveRDS(sinai_i_model_train_best, "../saved/sinai_i_model_train_best.Rds")
saveRDS(roc_df_i, "../saved/roc_df_i.Rds")
saveRDS(roc_curve_ci_i, "../saved/roc_curve_ci_i.Rds")

}

sinai_i_model_train_best = readRDS("../saved/sinai_i_model_train_best.Rds")
roc_df_i = readRDS("../saved/roc_df_i.Rds")
roc_curve_ci_i = readRDS("../saved/roc_curve_ci_i.Rds")

# build ROC curve again (because I didn't save it the first time!)
roc_curve_all_i <- pROC::roc(sinai_i_model_train_best$obs, sinai_i_model_train_best$IBD, levels = levels(sinai_i_model_train_best$obs))

# plot Using ggplot2
ggplot() +  
  # 95% Confidence Interval as a shaded region
  geom_ribbon(data = roc_curve_ci_i, aes(x = rev(spec), ymin = lower, ymax = upper), 
              fill = "blue", alpha = 0.2) +
  # ROC Curve
  geom_line(data = roc_df_i, aes(x = 1-specificity, y = sensitivity), color = "black", size = 1) +

  # Add diagonal reference line (random classifier)
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray") +
  # Labels and theme
  labs(title = paste("Ileum (Non-inflamed)\nROC Curve with 95% CI", " (AUC: ", round(roc_curve_all_i$auc, digits=3), ")", sep=""), x = "1 - Specificity", y = "Sensitivity") +
  theme_bw()

```


```{r, is_i_model_useful, echo=F, results=F, plot=T, warning=F, message=F}

# extract mean predictions and link to inflammation status
sinai_i_model_mean = sinai_i_model_train_best %>%
  group_by(rowIndex) %>%
  mutate(mean.prob = mean(IBD)) %>%
  dplyr::select(rowIndex, mean.prob) %>% distinct() %>% data.frame() %>%
  arrange(rowIndex) %>%
  mutate(Patient = rownames(sinai_joint_i))

sinai_i_model_mean = merge(sinai_i_model_mean,
                             subset(sinai_joint_meta, Location == "Rectum"), by="Patient")

sinai_i_model_mean$Disease = factor(sinai_i_model_mean$Disease, levels=c("Control", "CD", "UC"))
sinai_i_model_mean$Status = factor(sinai_i_model_mean$Status, levels=c("NonI", "I"))

sinai_i_model_mean$Status = factor(ifelse(sinai_i_model_mean$Status == "NonI", "Non-inflamed", "Inflamed"), levels=c("Non-inflamed", "Inflamed"))

ggplot(sinai_i_model_mean,
       aes(x=Disease, y=mean.prob))+
  geom_violin(draw_quantiles= 0.5) +
  ggbeeswarm::geom_beeswarm(shape=21, aes(fill = Disease), color="black")+
  geom_hline(yintercept=0.5, linetype=2, alpha=0.5)+
  theme_minimal()+theme(legend.position="none")+
  facet_wrap(~Status, scales="free_x")+
  labs(x="Disease", y="Mean Probability of IBD", title="Non-Inflamed Ileum Model")

```

This ileum-only model appears to perform better at predicting CD, which makes sense because inflammation in ileal tissue is generally specific to CD.

Now, let's build an ensemble model using a caret extension.

```{r, ensemble, echo=F, results=F, plot=T, warning=FALSE, message=FALSE}

cached = T

if(cached == F){
  
set.seed(25)
sinai_ensemble_model <- caretEnsemble::caretEnsemble(
  list(sinai_r_model_train,
       sinai_i_model_train),
  trControl = trainControl(
    method = "cv", 
    savePredictions = TRUE,
    classProbs = TRUE,
    summaryFunction = caret::twoClassSummary,
    sampling = "down"),
  metric = "ROC")

# select the final model
sinai_ensemble_model_train_best = sinai_ensemble_model$ens_model

# build ROC curve
roc_curve_all_ensemble <- pROC::roc(sinai_ensemble_model_train_best$pred$obs, sinai_ensemble_model_train_best$pred$IBD, levels = levels(sinai_ensemble_model_train_best$pred$obs))
# convert ROC Curve to Data Frame
roc_df_ensemble <- data.frame(
  specificity = (roc_curve_all_ensemble$specificities),
  sensitivity = roc_curve_all_ensemble$sensitivities
) %>% arrange((sensitivity))

# calculate 95% confidence interval
roc_curve_ci_ensemble <- pROC::ci.se(roc_curve_all_ensemble, specificities = seq(0, 1, l = 100), boot.n = 2000)
roc_curve_ci_ensemble = data.frame(roc_curve_ci_ensemble)
colnames(roc_curve_ci_ensemble) = c("lower", "mean", "upper")
roc_curve_ci_ensemble$spec = seq(0, 1, l = 100)

saveRDS(sinai_ensemble_model_train_best, "../saved/sinai_ensemble_model_train_best.Rds")
saveRDS(roc_df_ensemble, "../saved/roc_df_ensemble.Rds")
saveRDS(roc_curve_ci_ensemble, "../saved/roc_curve_ci_ensemble.Rds")

}

sinai_ensemble_model_train_best = readRDS("../saved/sinai_ensemble_model_train_best.Rds")
roc_df_ensemble = readRDS("../saved/roc_df_ensemble.Rds")
roc_curve_ci_ensemble = readRDS("../saved/roc_curve_ci_ensemble.Rds")

# build ROC curve again (because I didn't save it the first time!)
roc_curve_all_ensemble <- pROC::roc(sinai_ensemble_model_train_best$pred$obs, sinai_ensemble_model_train_best$pred$IBD, levels = levels(sinai_ensemble_model_train_best$pred$obs))

# plot Using ggplot2
ggplot() +  
  # 95% Confidence Interval as a shaded region
  geom_ribbon(data = roc_curve_ci_ensemble, aes(x = rev(spec), ymin = lower, ymax = upper), 
              fill = "blue", alpha = 0.2) +
  # ROC Curve
  geom_line(data = roc_df_ensemble, aes(x = 1-specificity, y = sensitivity), color = "black", size = 1) +

  # Add diagonal reference line (random classifier)
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray") +
  # Labels and theme
  labs(title = paste("Ensemble (Non-inflamed)\nROC Curve with 95% CI", " (AUC: ", round(roc_curve_all_ensemble$auc, digits=3), ")", sep=""), x = "1 - Specificity", y = "Sensitivity") +
  theme_bw()

```

Does this classifier work more equally for CD and UC?

```{r, is_ensemble_useful, echo=F, results=F, plot=T, warning=F, message=F}

# extract mean predictions and link to inflammation status
sinai_ensemble_model_mean = sinai_ensemble_model_train_best$pred %>%
  group_by(rowIndex) %>%
  mutate(mean.prob = mean(IBD)) %>%
  dplyr::select(rowIndex, mean.prob) %>% distinct() %>% data.frame() %>%
  arrange(rowIndex) %>%
  mutate(Patient = rownames(sinai_joint_i))

sinai_ensemble_model_mean = merge(sinai_ensemble_model_mean,
                             subset(sinai_joint_meta, Location == "Rectum"), by="Patient")

sinai_ensemble_model_mean$Disease = factor(sinai_ensemble_model_mean$Disease, levels=c("Control", "CD", "UC"))
sinai_ensemble_model_mean$Status = factor(sinai_ensemble_model_mean$Status, levels=c("NonI", "I"))

sinai_ensemble_model_mean$Status = factor(ifelse(sinai_ensemble_model_mean$Status == "NonI", "Non-inflamed", "Inflamed"), levels=c("Non-inflamed", "Inflamed"))

ggplot(sinai_ensemble_model_mean,
       aes(x=Disease, y=mean.prob))+
  geom_violin(draw_quantiles= 0.5) +
  ggbeeswarm::geom_beeswarm(shape=21, aes(fill = Disease), color="black")+
  geom_hline(yintercept=0.5, linetype=2, alpha=0.5)+
  theme_minimal()+theme(legend.position="none")+
  facet_wrap(~Status, scales="free_x")+
  labs(x="Disease", y="Mean Probability of IBD", title="Non-Inflamed Ensemble Model")

```

We see the model is near equally sensitive for CD and UC. In other words, it's doing what we wanted this sort of ensemble to do.

# Conclusions

There are several reasons why the transcriptome of *non-inflamed* intestinal biopsies could be different between IBD and non-IBD controls. 

For instance, IBD is treated with immunomodulatory drugs, which can leave a transcriptomic footprint in the tissue, inflamed or not.

Or, longstanding IBD in remission may have a persistent transcriptomic signature (including what the authors call "molecular inflammation"), despite already having been diagnosed and treated in the past.

Therefore, there is limited utility in this particular model without having included new-onset, treatment-naive patients.



