---
title: "Predicting Disease (Microbiome"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library("dplyr")
library("ggplot2")

cache = T

```

# **Validating** the analysis

[The Segata Lab](http://segatalab.cibio.unitn.it/tools/metaml/index.html) published a [study](https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1004977) in 2016 employing machine learning on microbiomes to predict various diseases, including liver cirrhosis, colorectal cancer, IBD, obesity, and type 2 diabetes.

The stool microbiome [dataset](https://github.com/SegataLab/metaml/tree/master/data) is publicly available.

Below, we validate their findings and extend the analysis to employ **Multi-Class Classification** with **downsampling**.

## Loading data

Let's load the data and see what type of samples are included:

```{r load_data, echo=F, results=T, plot=T, warning=F, message=F}

# load stool microbiome data
aml_data = read.csv("../metaaml_validation/abundance_stoolsubset.txt", sep="\t")
# standardize sample name
colnames(aml_data) <- c("dataset_name", aml_data[1,-1])

# extract metadata from this file
aml_meta = aml_data[1:210,] %>% t() %>% data.frame()
# make first row column names
colnames(aml_meta) = aml_meta[1,]
# and delete this row
aml_meta = aml_meta[-1,]
# clean up to necessary columns
aml_meta = aml_meta[,c("sampleID", "subjectID", "disease", "pubmedid")]
# keep distinct
aml_meta = distinct(aml_meta)
# check stats
aml_meta[,c("disease")] %>% table()


```

This is a little messy, so let's clean up the samples into clean categories.

## Metadata processing

```{r metadata_processing, echo=F, results=T, plot=T, warning=F, message=F}

# subset to confirmed categories

aml_meta = subset(aml_meta, disease %in% c("cirrhosis","ibd_crohn_disease","ibd_ulcerative_colitis", "cancer", "n", "obesity", "t2d"))

# rename diseases
aml_meta$disease = ifelse(aml_meta$disease == "n", "none",
                          ifelse(aml_meta$disease == "ibd_crohn_disease", "IBD",
                                 ifelse(aml_meta$disease == "ibd_ulcerative_colitis",
                                        "IBD",
                                 aml_meta$disease)))

aml_meta[,c("disease")] %>% table()


```

A small note, but it's unclear which categories of samples contributed to the 981 control samples listed in the original study. I only see 944. As we see later, this slight decrease becomes amplified on a pursuit towards high quality data.

Let's also check if there are sample duplicates:


```{r metadata_duplicates_1, echo=F, results=T, plot=T, warning=F, message=F}

# check if patient duplicates

aml_meta$subjectID %>% table() %>% data.frame() %>%
  ggplot()+
  ggridges::geom_density_ridges(aes(x=Freq, y=1),
                                  color="black", fill="white", alpha=0.2)+
  theme_minimal()+
  labs(title = "Sample Replicates",
       x = "Number of Samples per Subject",
       y="Frequency")

```

Uh oh. That's a notable number of pseudoreplicates. Which condition do they belong to?

```{r metadata_duplicates_2, echo=F, results=T, plot=T, warning=F, message=F}

aml_meta[,c("subjectID","disease")] %>% table() %>% data.frame() %>%
  group_by(disease) %>%
  summarize(sum(Freq>1))

```

Many of the controls were replicates, which we can spare. However, 66 / 173 of the Crohn's disease samples were replicates, which we'd obviously rather not spare. But we need to if we want to apply machine learning without blatantly overfitting. So, we'll take only the first entry per subject.

Here are the final tallies of samples after removing pseudoreplicates.

```{r metadata_duplicates_3, echo=F, results=T, plot=T, warning=F, message=F}

aml_meta = aml_meta %>%
  group_by(subjectID) %>%
  slice_max(sampleID, n=1) %>% data.frame()
aml_meta[,c("disease")] %>% table()

```

Alright. Now let's check the ratios of control to cases per study among studies of disease.

```{r metadata_duplicates_4, echo=F, results=T, plot=T, warning=F, message=F}

# aml_meta[,c("pubmedid", "disease")] %>% table()


# let's rename the studyID here
aml_meta$studyID = ifelse(aml_meta$pubmedid == "20203603", "IBD",
                          ifelse(aml_meta$pubmedid == "24997787", "IBD",
                          ifelse(aml_meta$pubmedid == "23023125", "T2D",
                                 ifelse(aml_meta$pubmedid == "23719380", "WT2D",
                                        ifelse(aml_meta$pubmedid == "23985870", "obesity",
                                               ifelse(aml_meta$pubmedid == "25079328", "cirrhosis",
                                                      ifelse(aml_meta$pubmedid == "25432777", "cancer",
                                                             aml_meta$pubmedid)))))))

aml_meta[,c("studyID", "disease")] %>% table()


```
Every study of disease also included approximately as many, or more control samples as case samples.

Now let's process the sequencing data.

## Data processing

Let's extract out the species data, and check out the data distributions.

```{r data_processing_1, echo=F, results=T, plot=T, warning=F, message=F}

# subset to taxa data
aml_taxa = aml_data[-c(1:210),] %>% t() %>% data.frame()

# make first row colnames, and remove first row
colnames(aml_taxa) = aml_taxa[1,]
aml_taxa = aml_taxa[-1,]

# subset to only s__ annotations
aml_taxa = aml_taxa[,grepl("s__", colnames(aml_taxa))]
# remove strains
aml_taxa = aml_taxa[,!grepl("t__", colnames(aml_taxa))]
# clean names
colnames(aml_taxa) = gsub(".*s__", "", colnames(aml_taxa))

# make values numeric
aml_taxa = data.matrix(aml_taxa)

# 826 species

print(paste(ncol(aml_taxa), "species remain.", sep=" "))

if(cache == F){
# check distributions
aml_taxa_dist_plot = ggplot(reshape2::melt(as.matrix(aml_taxa)))+
  ggridges::geom_density_ridges(aes(x=value, group=Var2, y=1),
                                  color="black", fill="white", alpha=0.2)+
  theme_minimal()+
  labs(x="Count", y="Frequency")
saveRDS(aml_taxa_dist_plot, "../saved/aml_taxa_dist_plot.Rds")
}

aml_taxa_dist_plot <- readRDS("../saved/aml_taxa_dist_plot.Rds")

aml_taxa_dist_plot

```

We should convert to relative abundance, log transform, and add a pseudocount.

```{r data_processing_2, echo=F, results=T, plot=T, warning=F, message=F}

# pseudocount (min count was 1)
aml_taxa_pseudo = min(aml_taxa[aml_taxa!=0])/2

# add pseudocount
aml_taxa = aml_taxa+aml_taxa_pseudo

# convert to relative abundance
aml_taxa = sweep(aml_taxa, 1, rowSums(aml_taxa), FUN = "/")

# log transform
aml_taxa = log(aml_taxa, base=2)

if(cache == F){
# check distributions
aml_taxa_dist_plot_2 <- ggplot(reshape2::melt(as.matrix(aml_taxa)))+
  ggridges::geom_density_ridges(aes(x=value, group=Var2, y=1),
                                  color="black", fill="white", alpha=0.2)+
  theme_minimal()+
  labs(x="Relative Abundance", y="Frequency")
  
saveRDS(aml_taxa_dist_plot_2, "../saved/aml_taxa_dist_plot_2.Rds")
}

aml_taxa_dist_plot_2 <- readRDS("../saved/aml_taxa_dist_plot_2.Rds")

aml_taxa_dist_plot_2
```

Log transforming improved the distributions to approximately normal, but still right skewed. Nonetheless, it's appropriate to continue on to building models.

## Model building

The authors approach the machine learning task in a variety of ways:
*Our multi-level validation strategy includes the assessment of microbiome models on single cohorts, across stages of the same study, across different studies, and across target outcomes and conditions*

I'm interested in validating [Fig 2](https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1004977#pcbi.1004977.ref032), where they calculte a variety of metrics including AUC, for each of the datasets separately. Of note, they evaluate both Random Forest and SVM models. For species-based models, they employed a radial basis kernel.

Since they also mention evaluating elastic net, let's go ahead and additional models.

They also compared their models against null datasets, where the class label was randomly shuffled.

So, we'll do the following:

- For each of the 6 diseases/datasets:

- Evaluate 5 machine learning algorithms,

- Building 15 models,

- Using real and null data,

- And extract the internal validation AUCs.

```{r model_building_1, echo=F, results=F, plot=T, warning=F, message=F}

# data
# aml_taxa
# aml_meta c("sampleID", "pubmedid", "disease")

datasets = c("IBD", "cirrhosis", "cancer", "T2D", "WT2D", "obesity")
ml.models = c("glmnet","ranger","svmLinear","svmRadial","xgbTree")

if(cache == F){
  
# loop through each dataset for SVM and ranger
set.seed(25)
t1 <- Sys.time()
aml_model_loop = 
  do.call(rbind, lapply(c("real", "null"), function(type){
  do.call(rbind, lapply(1:15,function(seed){
  do.call(rbind, lapply(ml.models, function(algo){
        print(paste(type, " ", seed, " ", algo, " ", sep=""))

  do.call(rbind, parallel::mclapply(datasets, function(x){
   #  x = c("IBD", "cirrhosis", "cancer", "T2D", "WT2D", "obesity")[6]
  # subset metadata
  metadata.subset = aml_meta %>% subset(studyID == x) %>% select(sampleID, disease) %>% data.frame()
  metadata.subset$disease = ifelse(metadata.subset$disease == "none", "none", "disease")
  # subset data
  data.subset = aml_taxa[make.names(metadata.subset$sampleID),] %>% data.frame()
  # append class label
  data.subset$disease = metadata.subset[match(rownames(data.subset), make.names(metadata.subset$sampleID)),]$disease
  # interestingly, we can just remove the NA because they are duplicated (note the .1 added to   rowname)
  data.subset = subset(data.subset, !is.na(disease))
  data.subset$disease = as.factor(data.subset$disease)
 #  table(data.subset$disease)
  
  # conditionally shuffle class labels
    if(type == "null"){
    set.seed(seed)
    data.subset$disease = sample(data.subset$disease, replace=F)
    }
  
    # build model with downsampling
# conditionally modify glmnet tuning
  if(algo == "glmnet"){
      set.seed(seed)
    aml_train = caret::train(
  disease ~ .,
  data = data.subset,
  trControl = caret::trainControl(method="cv", 
                                  number = 5,
                                  savePredictions = TRUE,
                                  classProbs = TRUE,
                                  summaryFunction = caret::twoClassSummary,
                                  sampling = "down"),
  tuneGrid = expand.grid(
  alpha = seq(0, 1, length = 10),
  lambda = 10^seq(2, -4, length = 100)),
  metric = "ROC",
  method = "glmnet",
  preProc = c("center", "scale")
)
}else{
  # build model with downsampling
  set.seed(seed)
  aml_train <- caret::train(
    disease ~ .,
    data = data.subset,
    trControl = caret::trainControl(method="cv", 
                                    number = 5,
                                    savePredictions = TRUE,
                                    classProbs = TRUE,
                                    summaryFunction = caret::twoClassSummary,
                                    sampling = "down"),
    metric = "ROC",
    method = algo,
    preProc = c("center", "scale")) %>% suppressWarnings()
}
  # extract AUC
  data.frame(AUC = max(aml_train$results$ROC),
             Sensitivity = subset(aml_train$results, ROC == max(aml_train$results$ROC))$Sens,
             Specificity = subset(aml_train$results, ROC == max(aml_train$results$ROC))$Spec,
             model = algo,
             data = x,
             iter = seed,
             datatype = type)
  }))
  }))
  }))
  }))
   t2 <- Sys.time()
 t2-t1 # 7 min for all
 
 saveRDS(aml_model_loop, "../saved/aml_model_loop.Rds")
}


 aml_model_loop =  readRDS("../saved/aml_model_loop.Rds")

 # apply statistics
 
aml_model_ttest = do.call(rbind, lapply(ml.models, function(algo){
  do.call(rbind, lapply(datasets, function(x){
    print(paste(algo, " ", x))
    data.subset = aml_model_loop %>% distinct() %>% subset(model == algo & data == x)
    ttest_auc_result = t.test(subset(data.subset, datatype == "real")$AUC,
           subset(data.subset, datatype == "null")$AUC)$p.value
    ttest_sens_result = t.test(subset(data.subset, datatype == "real")$Sensitivity,
           subset(data.subset, datatype == "null")$Sensitivity)$p.value
    ttest_spec_result = t.test(subset(data.subset, datatype == "real")$Specificity,
           subset(data.subset, datatype == "null")$Specificity)$p.value
    data.frame(model = algo,
               data = x,
               AUC = ttest_auc_result,
               Sensitivity = ttest_sens_result,
               Specificity = ttest_spec_result)
  }))
  })) %>% reshape2::melt() %>%
  mutate(padj = p.adjust(value, method="bonferroni")) %>%
  mutate(sig = ifelse(padj < 0.05, "*", NA))

# make long df
aml_model_loop_df = reshape::melt(aml_model_loop, 
                     id=c("model","data","iter","datatype"),
                     variable.name = c("AUC", "Sensitivity","Specificity"))
# extract median value
aml_model_loop_median = aml_model_loop_df %>%
  group_by(data, model, datatype,variable) %>%
  mutate(med.val = median(value)) %>% 
  dplyr::select(data,model,datatype,variable,med.val) %>% distinct()

ggplot()+
  geom_violin(data=subset(aml_model_loop_df, datatype=="null"), aes(x = model, y=value), fill="grey",color=NA, alpha=0.6)+
  geom_violin(data=subset(aml_model_loop_df, datatype=="real"), aes(x = model, y=value, fill=model), alpha=0.6)+
  geom_point(data=subset(aml_model_loop_median, datatype=="real"), 
             aes(x =  model, y=med.val, fill= model), shape=21, size=1.5)+
  geom_text(data=aml_model_ttest, aes(x=model, Inf, label=sig), vjust=1.2)+
  theme_bw()+theme(legend.position="none",
                           axis.text.x=element_text(angle=45,vjust=1, hjust=1))+
  facet_grid(variable ~ data, switch="y")+
  labs(x="Model", y="")

```

This plot shows the distributions of 15 repeats of using real (colored) data and null (grey) data to train each model. I ran a t-test on each comparison followed by a Bonferroni adjustment to control the false discovery rate. Significant values are indicated by a *.

Using the internal validation metrics, we can see that random forest (ranger) is generally superior to SVM across datasets. Elastic net (glmnet) is quite competitive with ranger.

We can also get a sense of how some diseases (e.g. obesity) are harder to predict than others (e.g. IBD).

Importantly, most of these values are either as good or better than those reported in the original study.

Let's consider these results validated and move on to an interesting technique rarely employed in the literature: multi-class machine learning.

# **Extending** the analysis

The it would be ideal to build ONE model that predict each of the various diseases.

We can formulate this in different ways, like a one-vs-all or several one-vs-one classifiers.

Here, I want to demonstrate how a multi-class classifier performs using pooled datasets.

## Multiclass Classifier

First, we'll build a multiclass random forest that handles all samples simultaneously:

```{r, multivariate_rf_loop, echo=F, results=F, plot=T, warning=F, message=F}

  # loop 15 times
  metadata.subset = aml_meta %>% select(sampleID, disease) %>% data.frame()
  # subset data
  data.subset = aml_taxa
  rownames(data.subset) = make.names(rownames(data.subset))
  data.subset = aml_taxa[make.names(metadata.subset$sampleID),] %>% data.frame()
  # append class label
  data.subset$disease = metadata.subset[match(rownames(data.subset), make.names(metadata.subset$sampleID)),]$disease
  # interestingly, we can just remove the NA because they are duplicated (note the .1 added to rowname)
  data.subset = subset(data.subset, !is.na(disease))
  data.subset$disease = as.factor(data.subset$disease)
  
if(cache == F){
aml_multi_rf_loop = do.call(rbind, lapply(1:15, function(seed){
  
  # run RF
  set.seed(seed)
  aml_multirf = ranger::ranger(disease ~ .,
                               data.subset,
                               probability = TRUE,
                               keep.inbag = TRUE) # critical to avoid overfitting
  
  # save dataframe with predictions, to calculate Confusion Matrix, AUC, Sens, and Spec later
    aml_multirf_predictions = data.frame(aml_multirf$predictions,
                                         iter = seed)
    
    aml_multirf_predictions
    }))

saveRDS(aml_multi_rf_loop, "../saved/aml_multi_rf_loop.Rds")
        
}

aml_multi_rf_loop = readRDS("../saved/aml_multi_rf_loop.Rds")

``` 


```{r, multivariate_rf_conf, echo=F, results=F, plot=T, warning=F, message=F}

# first, let's plot a confusion matrix

aml_multi_rf_conf = do.call(rbind, lapply(1:15, function(seed){
    
  aml_multi_subset = subset(aml_multi_rf_loop, iter == seed)
  aml_multi_subset$iter = NULL
  
  oob_class <- colnames(aml_multi_subset)[max.col(aml_multi_subset)]  # Convert probabilities to class labels

  oob_class <- factor(oob_class, levels = levels(data.subset$disease))

  # Compute confusion matrix based on OOB predictions
  conf_matrix <- confusionMatrix(oob_class, data.subset$disease)
  conf_matrix_df = conf_matrix$table %>% reshape2::melt()

  conf_matrix_df$iter = seed
  
  conf_matrix_df
}))

# add total true value per dataset
true_ref = data.subset$disease %>% table() %>% data.frame()
colnames(true_ref) = c("Reference", "Freq")

# collapse to mean value
aml_multi_rf_conf_df = aml_multi_rf_conf %>%
  merge(true_ref, by="Reference") %>%
  group_by(Prediction, Reference) %>%
  # calculate mean pred and sd
  mutate(mean.val = mean(value), sd.val = sd(value)) %>%
  # create variable with % accuracy
  group_by(Reference) %>%
  mutate(mean.perc = mean.val / Freq) %>%
  dplyr::select(Prediction, Reference, mean.val, mean.perc, sd.val) %>% distinct()

# plot confusion matrix
aml_multi_rf_conf_plot = ggplot(aml_multi_rf_conf_df,
       aes(x=Reference, y = Prediction))+
  geom_tile(color="white", aes(fill = mean.perc*100))+
  geom_text(aes(label = paste0(round(mean.val, digits=0), "\n(",round(mean.perc*100, digits=0),"%)", sep="")),
            color = "white")+
  scale_fill_gradient2(low = scales::muted("blue"),
                       mid = "white", midpoint = 50,
                       high = scales::muted("red"))+
  theme_minimal()+
  labs(x="True", y="Predicted", fill="Mean\nPercentage\nPrediction")

aml_multi_rf_conf_plot

``` 

This confusion matrix illustrates a general low accuracy in predicting each disease. That is, the values in parentheses along the diagonal tiles are quite low for most diseases. The model is predicting most microbiomes belong to healthy individuals.

I suspect the massive class imbalance (e.g. n=665 "none" vs n=48 "cancer") is causing the model to bias its prediction towards the overrepresented class. 

What do the ROC curves look like?

```{r, multivariate_rf_auc, echo=F, results=F, plot=T, warning=F, message=F}

if(cache == F){
aml_multi_rf_auc_loop = do.call(rbind, lapply(1:15, function(seed){
  
  # run RF
  set.seed(seed)
  aml_multirf = ranger::ranger(disease ~ .,
                               data.subset,
                               probability = TRUE,
                               keep.inbag = TRUE) # critical to avoid overfitting
  
  aml_multirf_roc <- pROC::multiclass.roc(data.subset$disease, 
                                          aml_multirf$predictions, 
                                          levels = levels(data.subset$disease))
  
  # loop through each disease to calculate Sens and Spec
  # initialize dataframe
  roc_data = data.frame()
  for(dis in levels(data.subset$disease)){
    # extract disease-specific ROC
    labels = as.numeric(data.subset$disease == dis)
    roc_obj = pROC::roc(labels, data.frame(aml_multirf$predictions)[[dis]])
    # make dataframe
    roc_obj_df = data.frame(
      TPR = rev(roc_obj$sensitivities),
      FPR = rev(1-roc_obj$specificities),
      Class = dis
    )
    # append previous data
    roc_data = rbind(roc_data, roc_obj_df)
  }
    # add seed
  roc_data$iter == seed
  # add overall AUC
  roc_data$auc = aml_multirf_roc$auc
  
  roc_data
  
}))

# compute mean and CI
aml_multi_rf_auc_loop_summary <- aml_multi_rf_auc_loop %>%
  group_by(Class, TPR) %>%
  summarise(
    Mean_Sensitivity = mean(FPR),
    Lower_CI = quantile(FPR, 0.025),
    Upper_CI = quantile(FPR, 0.975),
    .groups = "drop"
  ) %>%
  group_by(Class) %>%
  arrange(TPR) %>%
  mutate(Mean_Sensitivity = cummax(Mean_Sensitivity)) %>%
  ungroup()
  
# compute mean AUC
mean_auc = mean(aml_multi_rf_auc_loop$auc)

# plot Using ggplot2
aml_multi_rf_auc_plot = ggplot() +  
  # mean ROC Curve
  geom_line(data = aml_multi_rf_auc_loop_summary, aes(x = Mean_Sensitivity, y = TPR, color=Class), size = 1) +
  # Add diagonal reference line (random classifier)
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray") +
  # Labels and theme
  labs(title = paste("Multiclass AUC", " (", round(mean_auc, digits=3), ")", sep=""), x = "1 - Specificity", y = "Sensitivity", color="Disease") +
  theme_minimal()
aml_multi_rf_auc_plot

saveRDS(aml_multi_rf_auc_plot, "../saved/aml_multi_rf_auc_plot.Rds")
        
}

aml_multi_rf_auc_plot = readRDS("../saved/aml_multi_rf_auc_plot.Rds")

aml_multi_rf_auc_plot

``` 

This ROC curve seems to suggest all diseases are being predicted reasonably well. However, "none" is predicted the poorest, likely because of the high degree of false positives observed in the confusion matrix.

## Downsample

Let's downsample these data down to the smallest dataset (n=48). I suspect this will sacrifice accuracy, but improve the false positive rate.

```{r, multivariate_rf_loop_balanced, echo=F, results=F, plot=T, warning=F, message=F}

  # loop 15 times
  metadata.subset = aml_meta %>% select(sampleID, disease) %>% data.frame()
  # subset data
  data.subset = aml_taxa
  rownames(data.subset) = make.names(rownames(data.subset))
  data.subset = aml_taxa[make.names(metadata.subset$sampleID),] %>% data.frame()
  # append class label
  data.subset$disease = metadata.subset[match(rownames(data.subset), make.names(metadata.subset$sampleID)),]$disease
  # interestingly, we can just remove the NA because they are duplicated (note the .1 added to rowname)
  data.subset = subset(data.subset, !is.na(disease))
  data.subset$disease = as.factor(data.subset$disease)
  
if(cache == F){
  
aml_multi_rf_loop_sample = do.call(rbind, lapply(1:15, function(seed){
  
  # downsample
  set.seed(seed)
  data.subset.sample = data.subset %>%
    group_by(disease) %>%
    slice_sample(n=48)
  
  # run RF
  set.seed(seed)
  aml_multirf_sample = ranger::ranger(disease ~ .,
                               data.subset.sample,
                               probability = TRUE,
                               keep.inbag = TRUE) # critical to avoid overfitting
  
    # save dataframe with predictions, to calculate Confusion Matrix, AUC, Sens, and Spec later
    aml_multirf_predictions_sample = data.frame(aml_multirf_sample$predictions,
                                         iter = seed)
    
    # add truth
    aml_multirf_predictions_sample$true = data.subset.sample$disease
    # add AUC for plotting
    aml_multirf_predictions_sample$auc = aml_multirf_sample$auc
    aml_multirf_predictions_sample
    
    }))

saveRDS(aml_multi_rf_loop_sample, "../saved/aml_multi_rf_loop_sample.Rds")
        
}

aml_multi_rf_loop_sample = readRDS("../saved/aml_multi_rf_loop_sample.Rds")

``` 


```{r, multivariate_rf_conf_balanced, echo=F, results=F, plot=T, warning=F, message=F}

# first, let's plot a confusion matrix

aml_multi_rf_conf_sample = do.call(rbind, lapply(1:15, function(seed){
    
  aml_multi_subset_sample = subset(aml_multi_rf_loop_sample, iter == seed)
  aml_multi_subset_sample$iter = NULL
  # extract out labels
  sample_labels = aml_multi_subset_sample$true
  aml_multi_subset_sample$true = NULL
  
  oob_class_sample <- colnames(aml_multi_subset_sample)[max.col(aml_multi_subset_sample)]  # Convert probabilities to class labels

  oob_class_sample <- factor(oob_class_sample, levels = levels(sample_labels))

  # Compute confusion matrix based on OOB predictions
  conf_matrix_sample <- confusionMatrix(oob_class_sample, sample_labels)
  conf_matrix_sample_df = conf_matrix_sample$table %>% reshape2::melt()

  conf_matrix_sample_df$iter = seed
  
  conf_matrix_sample_df
}))

# collapse to mean value
aml_multi_rf_conf_sample_df = aml_multi_rf_conf_sample %>%
  group_by(Prediction, Reference) %>%
  # calculate mean pred and sd
  mutate(mean.val = mean(value), sd.val = sd(value)) %>%
  # create variable with % accuracy
  group_by(Reference) %>%
  mutate(mean.perc = mean.val / 48) %>%
  dplyr::select(Prediction, Reference, mean.val, mean.perc, sd.val) %>% distinct()

aml_multi_rf_conf_sample_plot = ggplot(aml_multi_rf_conf_sample_df,
       aes(x=Reference, y = Prediction))+
  geom_tile(color="white", aes(fill = mean.perc*100))+
  geom_text(aes(label = paste0(round(mean.val, digits=0), "\n(",round(mean.perc*100, digits=0),"%)", sep="")),
            color = "white")+
  scale_fill_gradient2(low = scales::muted("blue"),
                       mid = "white", midpoint = 50,
                       high = scales::muted("red"))+
  theme_minimal()+
  labs(x="True", y="Predicted", fill="Mean\nPercentage\nPrediction")

aml_multi_rf_conf_sample_plot

``` 

This looks significantly more like a confusion matrix. The values in parentheses along the diagonal indicate the accuracy of each disease being predicted correctly.

Let's see the ROC curve:

```{r, multivariate_rf_auc_balanced, echo=F, results=F, plot=T, warning=F, message=F}

  # loop 15 times
  metadata.subset = aml_meta %>% select(sampleID, disease) %>% data.frame()
  # subset data
  data.subset = aml_taxa
  rownames(data.subset) = make.names(rownames(data.subset))
  data.subset = aml_taxa[make.names(metadata.subset$sampleID),] %>% data.frame()
  # append class label
  data.subset$disease = metadata.subset[match(rownames(data.subset), make.names(metadata.subset$sampleID)),]$disease
  # interestingly, we can just remove the NA because they are duplicated (note the .1 added to rowname)
  data.subset = subset(data.subset, !is.na(disease))
  data.subset$disease = as.factor(data.subset$disease)
  
if(cache == F){
  
aml_multi_rf_auc_sample = do.call(rbind, lapply(1:15, function(seed){
  
  # downsample
  set.seed(seed)
  data.subset.sample = data.subset %>%
    group_by(disease) %>%
    slice_sample(n=48)
  
  # run RF
  set.seed(seed)
  aml_multirf_sample = ranger::ranger(disease ~ .,
                               data.subset.sample,
                               probability = TRUE,
                               keep.inbag = TRUE) # critical to avoid overfitting
  
    # save dataframe with predictions, to calculate Confusion Matrix, AUC, Sens, and Spec later
    aml_multirf_predictions_sample = data.frame(aml_multirf_sample$predictions,
                                         iter = seed)
    
    # add truth
    aml_multirf_predictions_sample$true = data.subset.sample$disease
    # add AUC for plotting
    aml_multirf_predictions_sample$auc = aml_multirf_sample$auc
    aml_multirf_predictions_sample
    
  # loop through each disease to calculate Sens and Spec
  # initialize dataframe
  roc_data_sample = data.frame()
  for(dis in levels(data.subset$disease)){
    # extract disease-specific ROC
    sample_labels = as.numeric(aml_multirf_predictions_sample$true == dis)
    roc_obj_sample = pROC::roc(sample_labels, data.frame(aml_multirf_sample$predictions)[[dis]])
    # make dataframe
    roc_obj_sample_df = data.frame(
      TPR = rev(roc_obj_sample$sensitivities),
      FPR = rev(1-roc_obj_sample$specificities),
      Class = dis
    )
    # append previous data
    roc_data_sample = rbind(roc_data_sample, roc_obj_sample_df)
  }
    # add seed
  roc_data_sample$iter == seed
  # add overall AUC
  roc_data_sample$auc = roc_obj_sample$auc
  
  roc_data_sample
  
}))

# compute mean and CI
aml_multi_rf_auc_loop_summary_sample <- aml_multi_rf_auc_sample %>%
  group_by(Class, TPR) %>%
  summarise(
    Mean_Sensitivity = mean(FPR),
    Lower_CI = quantile(FPR, 0.025),
    Upper_CI = quantile(FPR, 0.975),
    .groups = "drop"
  ) %>%
  group_by(Class) %>%
  arrange(TPR) %>%
  mutate(Mean_Sensitivity = cummax(Mean_Sensitivity)) %>%
  ungroup()
  
# compute mean AUC
mean_auc_sample = mean(aml_multi_rf_auc_sample$auc)

# plot 
aml_multi_rf_auc_sample_plot = ggplot() +  
  # mean ROC Curve
  geom_line(data = aml_multi_rf_auc_loop_summary_sample, aes(x = Mean_Sensitivity, y = TPR, color=Class), size = 1) +
  # Add diagonal reference line (random classifier)
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray") +
  # Labels and theme
  labs(title = paste("Downsampled Multiclass AUC", " (", round(mean_auc_sample, digits=3), ")", sep=""), x = "1 - Specificity", y = "Sensitivity", color="Disease") +
  theme_minimal()
aml_multi_rf_auc_sample_plot

saveRDS(aml_multi_rf_auc_sample_plot, "../saved/aml_multi_rf_auc_sample_plot.Rds")
        
}
  

aml_multi_rf_auc_sample_plot = readRDS("../saved/aml_multi_rf_auc_sample_plot.Rds")

aml_multi_rf_auc_sample_plot

```

Once again, this ROC curve seems to suggest all diseases are being predicted at least reasonably well. We see that the AUC has shrunk by ~5%. We also see that the model performs most poorly with "None" and "T2D", in agreement with the confusion matrix.


# Conclusion

To conclude, we found that microbiome-based classifiers could predict disease quite well, particularly when control samples were confined to each study. 

When we built a model that tries to predict all samples together, we found that classification specificities and sensitivities suffered from large class imbalances. This illustrates the importance of balancing datasets when considering class-specific accuracies.

