---
title: "Predicting IBD using Metabolomes"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache=TRUE)
library("dplyr")
library("ggplot2")

cache = T

```


[The Huttenhower Lab](https://huttenhower.sph.harvard.edu/home/) conducted one of the largest and deepest characterizations of the IBD microbiome, incorporating metagenomics, metatranscriptomics, and metabolomics. [This study](https://www.nature.com/articles/s41564-018-0306-4) was published in Nature in 2019.

Their metabolomics data set has been reprocessed and uploaded by the [Borenstein lab](https://github.com/borenstein-lab), available [here](https://github.com/borenstein-lab/microbiome-metabolome-curated-data/tree/main/data/processed_data/FRANZOSA_IBD_2019). These data were incorporated in a [meta-analysis](https://www.nature.com/articles/s41522-022-00345-5) of gut microbiome-metabolome associations.

Towards the end of Franzosa's study, they use Random Forest to classify IBD and non-IBD microbiomes using metabolites, achieving an AUC of 0.92 using 5-fold cross-validation. Here, we attempt to validate this results.

# **Validating** the analysis

## Loading data

Let's load the metabolite feature table and check the dimensions:

```{r load_metabolite_counts, echo=F, results=T, plot=T, warning=F, message=F}

# load stool metabolome data
hmp_data = read.csv("../hmp_validation/mtb.tsv", sep="\t", 
                    check.names=F) # added to avoid losing special characters in colnames
rownames(hmp_data) = hmp_data$Sample
hmp_data$Sample = NULL

print(paste(dim(hmp_data)[1], " samples by ", dim(hmp_data)[2], " features", sep=""))

```

Since the metabolomics were performed untargeted, most of the features might not be annotated.

How many of the 8848 features were annotated?

```{r load_metabolite_annotated, echo=F, results=T, plot=T, warning=F, message=F}

# view annotated features
#colnames(hmp_data)[!grepl("\\: NA", colnames(hmp_data))] %>% length()

# count percentage of annotated features
print(paste0(sum(!grepl("\\: NA", colnames(hmp_data))), " features are annotated (",
            round(sum(!grepl("\\: NA", colnames(hmp_data))) / ncol((hmp_data)), digits=2)*100, "%)", sep=""))

```

Let's keep just these 466 features, which is a good feature space for machine learning, and renders a more interpretable conclusion when we look for important features.

```{r load_metabolite_map, echo=F, results=F, plot=T, warning=F, message=F} 

# molecule map
hmp_molecules = read.csv("../hmp_validation/mtb.map.tsv", sep="\t")
# likely not needed

```

Now let's check the sample data.

First, how many samples are in the original data set, and how many in the external validation sets?

```{r load_mapping, echo=F, results=T, plot=T, warning=F, message=F}

hmp_metadata = read.csv("../hmp_validation/metadata.tsv", sep="\t")
# head(hmp_metadata)

# extract out data set
hmp_metadata$data = ifelse(grepl("PRISM", hmp_metadata$Sample), "Training", "Validation")

hmp_metadata[,c("data", "Study.Group")] %>% table() %>% 
  data.frame() %>%
  `colnames<-`(c("dataset", "disease", "samples")) %>%
  arrange(dataset)

```

We'll be working with 34 control, 53 UC, and 68 CD samples to build the model, which is a bit of an imbalance worth correcting for. It's made worse by the fact we'll be pooling UC and CD together. The validation dataset was originally nicely balanced between n=20-23 per group, but these will also become imbalanced after pooling UC and CD.

Lastly, let's check for [pseudoreplicates](https://pdobrano25.github.io/ml_website/overfitcheck.html#Pseudoreplicates):

```{r check_pseudoreplicates, echo=F, results=F, plot=T, warning=F, message=F}

hmp_metadata[,c("Subject")] %>% table() %>% data.frame() %>%
  ggplot()+
  ggridges::geom_density_ridges(aes(x=Freq, y=1),
                                  color="black", fill="white", alpha=0.2)+
  theme_minimal()+
  labs(title = "Sample Replicates",
       x = "Number of Samples per Subject",
       y="Frequency")

```

Great! There are no repeat samplings of any subject.

## Data processing

Now we can check the feature distributions.

```{r check_distributions, echo=F, results=F, plot=T, warning=F, message=F}

# subset to annotated metabolites
hmp_data_an = hmp_data[,!grepl("\\: NA", colnames(hmp_data))]

hmp_met_dist_plot = ggplot(reshape2::melt(as.matrix(hmp_data_an)))+
  ggridges::geom_density_ridges(aes(x=value, group=Var2, y=1),
                                  color="black", fill="white", alpha=0.2)+
  theme_minimal()+
  labs(x="Abundance", y="Frequency")
hmp_met_dist_plot

```

Based on the x axis scale, I suspect these data are profoundly right-skewed and are asking to be log-transformed. Let's see what happens. We'll also need to add a pseudocount because of the 0's.

```{r check_distributions_log, echo=F, results=F, plot=T, warning=F, message=F}

# subset to annotated metabolites
hmp_pseudo = min(hmp_data_an[hmp_data_an!=0])/2

hmp_data_an = log(hmp_data_an+hmp_pseudo, base=2)

hmp_met_dist_plot_log = ggplot(reshape2::melt(as.matrix(hmp_data_an)))+
  ggridges::geom_density_ridges(aes(x=value, group=Var2, y=1),
                                  color="black", fill="white", alpha=0.2)+
  theme_minimal()+
  labs(x="Log Abundance", y="Frequency")
hmp_met_dist_plot_log

```

These look way better. Let's move on to building our models and validating them using internal CV.

## Model building

Franzosa employed 5x cross-validated random forest to predict IBD, achieving an AUC of 0.92. 
The authors specify the following:

*labels were randomly balanced before training and 100 trees were considered*

So, we'll employ downsampling and specify 100 trees per random forest.

```{r randomforest_internal, echo=F, results=F, plot=T, warning=F, message=F}

# finalize input dataset

# append class label
hmp_data_an$disease = hmp_metadata[match(rownames(hmp_data_an), hmp_metadata$Sample),]$Study.Group
# convert CD and UC to IBD
hmp_data_an$disease = ifelse(hmp_data_an$disease == "Control", "Control", "IBD")
# make a factor
hmp_data_an$disease = factor(hmp_data_an$disease, levels=c("Control", "IBD"))

hmp_internal = hmp_data_an[subset(hmp_metadata, data=="Training")$Sample,]

# build one random forest model

if(cache == F){
  
set.seed(25)
hmp_train = caret::train(
  disease ~ .,
  data = hmp_internal,
  trControl = caret::trainControl(method="cv", 
                                  number = 5,
                                  savePredictions = TRUE,
                                  classProbs = TRUE,
                                  summaryFunction = caret::twoClassSummary,
                                  sampling = "down"),
  num.trees = 100,
  importance = 'permutation',
  metric = "ROC",
  method = "ranger",
  preProc = c("center", "scale")
)
saveRDS(hmp_train, "../saved/hmp_train.Rds")
}

hmp_train = readRDS("../saved/hmp_train.Rds")

```

```{r randomforest_internal_aucs,  echo=F, results=F, plot=T, warning=F, message=F}

# extract AUC and ROC curve for internal

# select the final model
hmp_train_best = subset(hmp_train$pred, 
                        mtry == hmp_train$bestTune$mtry & 
                        min.node.size == hmp_train$bestTune$min.node.size &
                        splitrule == as.character(hmp_train$bestTune$splitrule))

# build ROC curve
hmp_train_roc <- pROC::roc(hmp_train_best$obs, hmp_train_best$IBD, levels = levels(hmp_train_best$obs))
# convert ROC Curve to Data Frame
hmp_train_roc_df <- data.frame(
  specificity = (hmp_train_roc$specificities),
  sensitivity = hmp_train_roc$sensitivities
) %>% arrange((sensitivity))

# calculate 95% confidence interval
hmp_train_ci <- pROC::ci.se(hmp_train_roc, specificities = seq(0, 1, l = 100), boot.n = 2000)
hmp_train_ci = data.frame(hmp_train_ci)
colnames(hmp_train_ci) = c("lower", "mean", "upper")
hmp_train_ci$spec = seq(0, 1, l = 100)

# plot using ggplot2
hmp_train_auc_plot = ggplot() +  
  # 95% Confidence Interval as a shaded region
  geom_ribbon(data = hmp_train_ci, aes(x = rev(spec), ymin = lower, ymax = upper), 
              fill = "blue", alpha = 0.2) +
  # ROC Curve
  geom_line(data = hmp_train_roc_df, aes(x = 1-specificity, y = sensitivity), color = "black", size = 1) +

  # Add diagonal reference line (random classifier)
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray") +
  # Labels and theme
  labs(title = paste("ROC Curve with 95% CI\nTraining (AUC: ", round(hmp_train_roc$auc, digits=3), ")", sep=""), x = "1 - Specificity", y = "Sensitivity") +
  theme_bw()

```

```{r randomforest_external_aucs,  echo=F, results=F, plot=T, warning=F, message=F}

# apply model to validation data

hmp_external_data =  hmp_data_an[subset(hmp_metadata, data!="Training")$Sample,]

# apply prediction
hmp_test_pred = predict(hmp_train,
                        hmp_external_data,
                        type="prob")

# build ROC curve
hmp_test_roc <- pROC::roc(hmp_external_data$disease, hmp_test_pred$IBD, levels = levels(hmp_external_data$disease))
# convert ROC Curve to Data Frame
hmp_test_roc_df <- data.frame(
  specificity = (hmp_test_roc$specificities),
  sensitivity = hmp_test_roc$sensitivities
) %>% arrange((sensitivity))

# calculate 95% confidence interval
hmp_test_ci <- pROC::ci.se(hmp_test_roc, specificities = seq(0, 1, l = 100), boot.n = 2000)
hmp_test_ci = data.frame(hmp_test_ci)
colnames(hmp_test_ci) = c("lower", "mean", "upper")
hmp_test_ci$spec = seq(0, 1, l = 100)

# plot using ggplot2
hmp_test_auc_plot = ggplot() +  
  # 95% Confidence Interval as a shaded region
  geom_ribbon(data = hmp_test_ci, aes(x = rev(spec), ymin = lower, ymax = upper), 
              fill = "blue", alpha = 0.2) +
  # ROC Curve
  geom_line(data = hmp_test_roc_df, aes(x = 1-specificity, y = sensitivity), color = "black", size = 1) +

  # Add diagonal reference line (random classifier)
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray") +
  # Labels and theme
  labs(title = paste("ROC Curve with 95% CI\nTesting (AUC: ", round(hmp_test_roc$auc, digits=3), ")", sep=""), x = "1 - Specificity", y = "Sensitivity") +
  theme_bw()

# plot both
hmp_train_auc_plot | hmp_test_auc_plot

```

These values (0.903 and 0.910) are close to those reported in the original study (0.92 and 0.89)

Now we can extend the analysis and apply additional models, under the hypothesis that random forest was suboptimal for this data set.

# **Extending** the analysis

We'll conduct 15 iterations of 5x CV and report AUCs from internal and external data sets. We'll continue to apply downsampling.

```{r, select_models, echo=F, results=FALSE, warning=FALSE, message=FALSE}
# select models
ml.models = c("gbm","glmnet","ranger","svmLinear","svmRadial","xgbTree")
```

```{r model_loop,  echo=F, results=F, plot=T, warning=F, message=F}

# loop through models
if(cache == F){
  
  t1 <- Sys.time()
hmp_auc_loop = do.call(rbind, lapply(1:15, function(seed){
  do.call(rbind, lapply(ml.models, function(algo){
    do.call(rbind, lapply(c("real", "null"), function(type){
       # run RF
      print(paste0(seed, " ", algo, " ", type))
      
      # check data type
      if(type == "real"){
      data_selected = hmp_internal
      }
      if(type == "null"){
        data_selected = hmp_internal
          set.seed(seed)
        data_selected$disease = sample(data_selected$disease)
      }
      
  set.seed(seed)
  hmp_model = caret::train(
    disease ~ .,
    data = data_selected,
    trControl = caret::trainControl(method="cv", 
                                  number = 5,
                                  savePredictions = TRUE,
                                  classProbs = TRUE,
                                  summaryFunction = caret::twoClassSummary,
                                  sampling = "down"),
    metric = "ROC",
    method = algo,
    preProc = c("center", "scale")
  ) %>% suppressWarnings() %>% suppressMessages()
  
  # calculate internal AUC
  hmp_internal_roc <- max(hmp_model$results$ROC)
  
  # apply model to external data
  hmp_model_test = predict(hmp_model, hmp_external_data, type="prob")

  # build ROC curve
  hmp_external_roc <- pROC::roc(hmp_external_data$disease, hmp_model_test$IBD, levels = levels(hmp_external_data$disease))
  
  # collate into data.frame
  data.frame(model = algo,
             iter = seed,
             internal_auc = hmp_internal_roc,
             external_auc = as.numeric(hmp_external_roc$auc),
             datatype = type)
  
}))
}))
}))
t2 <- Sys.time()
t2 - t1 # 1 hour
saveRDS(hmp_auc_loop, "../saved/hmp_auc_loop.Rds")

}

hmp_auc_loop = readRDS("../saved/hmp_auc_loop.Rds")


# make long df
hmp_auc_loop_df = reshape::melt(hmp_auc_loop, 
                     id=c("model","iter","datatype"),
                     variable.name = c("internal_auc", "external_auc"))


hmp_auc_ttest_df = do.call(rbind, lapply(ml.models, function(algo){
    print(paste(algo))
  
    data.subset = hmp_auc_loop_df %>% distinct() %>% subset(model == algo)
    
    ttest_internal_auc_result = t.test(subset(data.subset, datatype == "real" & variable == "internal_auc")$value,
           subset(data.subset, datatype == "null" & variable == "internal_auc")$value)$p.value
    
    ttest_external_auc_result = t.test(subset(data.subset, datatype == "real" & variable == "internal_auc")$value,
           subset(data.subset, datatype == "null" & variable == "internal_auc")$value)$p.value
    
    data.frame(model = algo,
               internal_auc = ttest_internal_auc_result,
               external_auc = ttest_external_auc_result)
    
  })) %>% reshape2::melt() %>%
  mutate(padj = p.adjust(value, method="bonferroni")) %>%
  mutate(sig = ifelse(padj < 0.05, "*", NA))


# extract median value
hmp_auc_loop_median = hmp_auc_loop_df %>%
  group_by(model, datatype,variable) %>%
  mutate(med.val = median(value)) %>% 
  dplyr::select(model,datatype,variable,med.val) %>% distinct()

ggplot()+
  geom_violin(data=subset(hmp_auc_loop_df, datatype=="null"), aes(x = model, y=value), fill="grey",color=NA, alpha=0.6)+
  geom_violin(data=subset(hmp_auc_loop_df, datatype=="real"), aes(x = model, y=value, fill=model), alpha=0.6)+
  geom_point(data=subset(hmp_auc_loop_median, datatype=="real"), 
             aes(x =  model, y=med.val, fill= model), shape=21, size=1.5)+
  geom_text(data=hmp_auc_ttest_df, aes(x=model, Inf, label=sig), vjust=1.2)+
  scale_y_continuous(breaks=seq(0.3,1, by=0.1))+
  theme_bw()+theme(legend.position="none",
                           axis.text.x=element_text(angle=45,vjust=1, hjust=1))+
  facet_wrap(~variable)+
  labs(x="Model", y="AUC")

```

All models are significantly superior to a model trained on null data.

Random forest is slightly suboptimal for the internal validation, but optimal for external validation, which suggests we've avoided overfitting.

## Feature importances

Despite testing additional models, we haven't learned anything new from the data.

Instead, we can assess feature importances to see which metabolites drive differences between IBD and controls.

Let's go back to the originally trained model and extract feature importances.

```{r, feature_importance, echo=F, results=F, plot=T, warning=F, message=F}

# extract feature importance
hmp_feature_imp = varImp(hmp_train)$importance %>%
  data.frame() %>%
  # add feature name
  mutate(feature = gsub("`", "", row.names(.))) %>%
  # clean feature name
  mutate(clean_feature = gsub(".*: ", "", feature))

# associate with IBD; calculate LFC
hmp_feature_lfc = do.call(rbind, lapply(1:ncol(hmp_data_an[colnames(hmp_data_an) != "disease"]), function(x){
  #x = 1
  data.subset = data.frame(feature = hmp_data_an[,x],
                           disease = hmp_data_an$disease)
  
  lm.output = lm(feature ~ disease, data.subset) %>% summary() %>% coef()
  data.frame(feature = colnames(hmp_data_an)[x],
             lfc = lm.output[2,1])
}))
hmp_feature_lfc

# merge
hmp_feature_imp = merge(hmp_feature_imp,
                        hmp_feature_lfc, by="feature")

# plot
ggplot(hmp_feature_imp %>% arrange(-Overall),
       aes(x=lfc, y=Overall))+
  geom_point(shape=21, aes(fill=lfc, size=Overall))+
  ggrepel::geom_label_repel(data = subset(hmp_feature_imp, Overall > 50),
                           aes(x=lfc, y=Overall, label=clean_feature), size=3)+
  scale_fill_gradient2(high=scales::muted("red"),
                       mid="white",
                       low=scales::muted("blue"))+
  labs(x="Log Fold Change", y="Feature Importance")+
  theme_minimal()+theme(legend.position="none")

```

Interestingly, the most important feature for discriminating IBD from controls was *urobilin*. Urobilin was found to be significantly decreased in IBD in another [cohort](https://pmc.ncbi.nlm.nih.gov/articles/PMC10359577/) in 2023, and validated [again](https://gut.bmj.com/content/73/8/1392) by a separate research group. Interestingly, reduced microbial bilirubin metabolism is a marker of IBD and previous antibiotic use ([source](https://www.nature.com/articles/s41564-023-01549-x)).

# Conclusions

To conclude, we were able to validate Franzosa's metabolome-based IBD classifier, seeing no significant improvement when using other machine learning models. By extracting feature importances, we found urobilin to be the top predictor of disease, which is highly accordant with the literature.


