---
title: "Predicting Metabolites using Metagenome Compositions"

---

```{r setup_mlp, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library("dplyr")
library("ggplot2")
cache=T
```


The [Borenstein lab](https://github.com/borenstein-lab) published a [study](https://microbiomejournal.biomedcentral.com/articles/10.1186/s40168-021-01149-z) seeking to predict metabolites in stool microbiomes using genus-level microbe abundances. This is not the first attempt at a research group trying to tackle this problem, [(for example)](https://www.nature.com/articles/s41467-019-10927-1), but the Borenstein group did great work collating several datasets made readily available for others to use.

At a simple level, Muller's study built machine learning (random forest) regression algorithms, using relative abundances of genera to predict the abundances of metabolites shared across datasets. They did additional analyses (e.g. random-effects models) to assess the generalizability of predictions across datasets. I am interested in simply seeing if bacterial abundances can predict metabolites, and whether additional models can outperform random forest. 

Below, I focus on only 2 of the datasets Muller studied: FRANZOSA and IHMP. Both of these datasets employ *whole genome shotgun sequencing* and *untargeted metabolomics* on IBD microbiomes. Below, I will:

- Build models using FRANZOSA's n=164 samples

- Evaluate the predictions on the IHMP's n=79 samples

- Test additional models that improve the predictive power.

- Extract feature importances and identify bacteria associated with metabolites

# **Validating** the analysis

## Loading metadata

First, we'll select the samples we want to work with. We already know there are [pseudoreplicates](https://pdobrano25.github.io/ml_website/overfitcheck.html#Pseudoreplicates) in the raw datasets, so let's select only unique IBD patients to work with:

```{r sample_picking, echo=F, results=T, plot=T, warning=F, message=F }

muller_meta = read.csv("../muller_validation/franzosa_metadata.tsv", sep="\t")
ihmp_meta = read.csv("../muller_validation/ihmp_metadata.tsv", sep="\t")

# combine them
muller_meta = rbind(ihmp_meta[,c("Dataset", "Sample", "Subject", "Study.Group")],
                    muller_meta[,c("Dataset", "Sample", "Subject", "Study.Group")]) %>% data.frame()
muller_meta$Study.Group = ifelse(muller_meta$Study.Group == "CD" | muller_meta$Study.Group == "UC", "IBD",
                                 "Control")
# subset to IBD patients
muller_meta = subset(muller_meta, Study.Group == "IBD")

# subset to unique IBD patients

muller_meta = muller_meta %>%
  group_by(Subject) %>%
  slice_max(Sample, n=1) %>% data.frame()

muller_meta[,c("Dataset", "Study.Group")] %>% table() %>% 
  data.frame() %>%
  `colnames<-`(c("dataset", "disease", "samples")) %>%
  arrange(dataset)

```

There are 164 unique subjects' samples in the Franzosa dataset, and 79 in the IHMP dataset.

## Loading metabolomic data

Now, we'll load both metabolomic datasets and restrict to annotated metabolites shared in each dataset.

```{r load_metabolite_data, eval=F, echo=F, results=T, plot=T, warning=F, message=F}

# note: we're going to hard code some of this, so we can apply filter and only load the processed data after
# we will need to write out a few things manually:

# load FRANZOSA data
franzosa_mbx = read.csv("../muller_validation/franzosa_mtb.tsv", sep="\t",
                                            check.names=F) # added to avoid losing special characters in colnames
dim(franzosa_mbx) # 220 x 8849

# load IHMP data
ihmp_mbx = read.csv("../muller_validation/ihmp_mtb.tsv", sep="\t",
                                            check.names=F) # added to avoid losing special characters in colnames

dim(ihmp_mbx) # 382 x 81868


# remove NA; tabulate annotated
franzosa_mbx = franzosa_mbx[,!grepl("\\: NA", colnames(franzosa_mbx))]
dim(franzosa_mbx)

ihmp_mbx = ihmp_mbx[,!grepl("\\__NA", colnames(ihmp_mbx))]
dim(ihmp_mbx)

# save these
saveRDS(franzosa_mbx, "../saved/franzosa_mbx_data.Rds")
saveRDS(ihmp_mbx, "../saved/ihmp_mbx_data.Rds")

```

```{r, process_metabolite_data, echo=F, results=T, plot=T, warning=F, message=F}
# remember to remove 1 feature because of Sample column
print("Franzosa contains 164 samples by 8,848 metabolites (467 annotated)")
print("IHMP contains 79 samples by 81,867 metabolites (597 annotated)")

# load RDS
franzosa_mbx = readRDS("../saved/franzosa_mbx_data.Rds")
ihmp_mbx = readRDS("../saved/ihmp_mbx_data.Rds")

# make rownames Sample
rownames(franzosa_mbx) = franzosa_mbx$Sample
rownames(ihmp_mbx) = ihmp_mbx$Sample
franzosa_mbx$Sample = NULL
ihmp_mbx$Sample = NULL

# subset to desired subjects
franzosa_mbx = franzosa_mbx[rownames(franzosa_mbx) %in% muller_meta$Sample,]
ihmp_mbx = ihmp_mbx[rownames(ihmp_mbx) %in% muller_meta$Sample,]

# check colname overlap
intersecting.mbx = intersect(gsub(".*\\: ", "", colnames(franzosa_mbx)), gsub(".*\\__", "", colnames(ihmp_mbx)))
print(paste0(length(intersecting.mbx), " intersecting metabolites"))

# subset to these compounds
franzosa_mbx = franzosa_mbx[,gsub(".*\\: ", "", colnames(franzosa_mbx)) %in% intersecting.mbx]
ihmp_mbx = ihmp_mbx[,gsub(".*\\__", "",colnames(ihmp_mbx)) %in% intersecting.mbx]

# clean colnames
colnames(franzosa_mbx) <- gsub(".*\\: ", "X__", colnames(franzosa_mbx))
colnames(ihmp_mbx) <- gsub(".*\\__", "X__", colnames(ihmp_mbx))

# good
```

We'll check the distributions of the datasets.

```{r distribution_check_mbx, echo=F, results=T, plot=T, warning=F, message=F }

# first, check each dataset's original distributions

franzosa_raw_metabolite_hist = ggplot(reshape2::melt(as.matrix(franzosa_mbx)))+
  ggridges::geom_density_ridges(aes(x=value, group=Var2, y=1),
                                  color="black", fill="white", alpha=0.2)+
  theme_minimal()+
  labs(x="Abundance", y="Frequency", title="Metabolomics\nFranzosa (Untransformed)")


ihmp_raw_metabolite_hist = ggplot(reshape2::melt(as.matrix(ihmp_mbx)))+
  ggridges::geom_density_ridges(aes(x=value, group=Var2, y=1),
                                  color="black", fill="white", alpha=0.2)+
  theme_minimal()+
  labs(x="Abundance", y="Frequency", title="Metabolomics\nIHMP (Untransformed)")



# next, log-transform, but select a pseudocount that can be applied to both datasets
# first, replace NA with 0
ihmp_mbx[is.na(ihmp_mbx)] = 0

franzosa_mbx_pseudo = min(franzosa_mbx[franzosa_mbx!=0])/2
ihmp_mbx_pseudo = min(ihmp_mbx[ihmp_mbx!=0])/2
# select smaller of the 2
muller_mbx_pseudo = min(c(franzosa_mbx_pseudo, ihmp_mbx_pseudo))

franzosa_mbx_log = log(franzosa_mbx+muller_mbx_pseudo, base=2)
ihmp_mbx_log = log(ihmp_mbx+muller_mbx_pseudo, base=2)


franzosa_log_metabolite_hist = ggplot(reshape2::melt(as.matrix(franzosa_mbx_log)))+
  ggridges::geom_density_ridges(aes(x=value, group=Var2, y=1),
                                  color="black", fill="white", alpha=0.2)+
  theme_minimal()+
  labs(x="Log Abundance", y="Frequency", title="Metabolomics\nFranzosa (Log-transformed)")

ihmp_log_metabolite_hist = ggplot(reshape2::melt(as.matrix(ihmp_mbx_log)))+
  ggridges::geom_density_ridges(aes(x=value, group=Var2, y=1),
                                  color="black", fill="white", alpha=0.2)+
  theme_minimal()+
  labs(x="Log Abundance", y="Frequency", title="Metabolomics\nIHMP (Log-transformed)")


# plot all
library("patchwork") %>% suppressMessages()
(franzosa_raw_metabolite_hist | ihmp_raw_metabolite_hist) / 
  (franzosa_log_metabolite_hist | ihmp_log_metabolite_hist)


```

Log-transforming the data with an added pseudocount had the effect we expected.

## Loading metagenomic data

Now we can load the metagenomic data, and as with the metabolites, we'll take the intersecting species.

```{r load_metagenomic_data, eval=F, echo=F, results=T, plot=T, warning=F, message=F }


# load FRANZOSA data (use read_tsv, otherwise it's too slow)
franzosa_mgx = readr::read_tsv("../muller_validation/franzosa_genera.tsv") %>% data.frame()
dim(franzosa_mgx) # 220 samples by 11,721 taxa
rownames(franzosa_mgx) = franzosa_mgx$Sample

# load IHMP data
ihmp_mgx = readr::read_tsv("../muller_validation/ihmp_genera.tsv") %>% data.frame()
dim(ihmp_mgx) # 382 x 9,695
rownames(ihmp_mgx) = ihmp_mgx$Sample

# subset to Genus (remove all other taxonomic annotations/levels)
franzosa_mgx = franzosa_mgx[,grepl("g__", colnames(franzosa_mgx))]
ihmp_mgx = ihmp_mgx[,grepl("g__", colnames(ihmp_mgx))]

# clean names
colnames(franzosa_mgx) = gsub(".*g__", "", colnames(franzosa_mgx))
colnames(ihmp_mgx) = gsub(".*g__", "", colnames(ihmp_mgx))

# subset to desired samples
franzosa_mgx = franzosa_mgx[rownames(franzosa_mgx) %in% muller_meta$Sample,]
ihmp_mgx = ihmp_mgx[rownames(ihmp_mgx) %in% muller_meta$Sample,]

# convert to %
franzosa_mgx = sweep(franzosa_mgx, 1, rowSums(franzosa_mgx), FUN = "/")
ihmp_mgx = sweep(ihmp_mgx, 1, rowSums(ihmp_mgx), FUN = "/")

# save these
saveRDS(franzosa_mgx, "../saved/franzosa_mgx_data.Rds")
saveRDS(ihmp_mgx, "../saved/ihmp_mgx_data.Rds")

```

We're going to need to apply some stringent prevalence and abundance filtering.

Let's see how many taxa remain across various thresholds.

```{r metagenomic_filtering_plot, echo=F, results=T, plot=T, warning=F, message=F }


# reload data 
franzosa_mgx = readRDS("../saved/franzosa_mgx_data.Rds")
ihmp_mgx = readRDS("../saved/ihmp_mgx_data.Rds")

# using franzosa, calculate prevalences across abundance thresholds
franzosa_mgx_prev = franzosa_mgx
if(cache==F){
franzosa_mgx_prev = 
  # select prevalence thresholds
  do.call(rbind, lapply(c(1,2,3,4,5,6,7,8,9,10), function(y){
    # select minimum abundance threshold
  do.call(rbind, lapply(c(0.000001, 0.00001, 0.0001, 0.001, 0.01), function(x){
  franzosa_mgx_prev[franzosa_mgx_prev >= x] <- 1
  franzosa_mgx_prev[franzosa_mgx_prev < x] <- 0
  prev = colSums(franzosa_mgx_prev) 
  # tabulate number of taxa remaining after abundance + prevalence thresholds
  data.frame(prev = sum(prev >= round(y*nrow(franzosa_mgx_prev)/100, digits=0)),
             abun_filt = x,
             prev_filt = y)
}))}))
saveRDS(franzosa_mgx_prev, "../saved/franzosa_mgx_prev.Rds")
}
franzosa_mgx_prev = readRDS( "../saved/franzosa_mgx_prev.Rds")

ggplot(franzosa_mgx_prev,
       aes(x=(abun_filt*100), y=(prev)))+
  geom_point(aes(fill=prev_filt), shape=21, size=3)+
  scale_fill_gradient2(low=scales::muted("blue"),
                       mid="white",midpoint=6,
                       high=scales::muted("red"))+
  scale_x_log10()+
  scale_y_log10()+
  theme_minimal()+
  labs(x="Abundance % Threshold",
       y="Taxa Remaining",
       fill="Prevalence\nThreshold\n(% samples)")

```

As we can see, we can preserve a little over 300 taxa if we apply a filter that keeps taxa "detected at 0.01% abundance in 10%+ samples"

Let's apply that filter and keep the intersecting (shared) taxa between datasets.

```{r metagenomic_filtering_apply, echo=F, results=T, plot=T, warning=F, message=F }

franzosa_mgx_filt_prev = franzosa_mgx

franzosa_mgx_filt_prev[franzosa_mgx_filt_prev >= 0.0001] <- 1
franzosa_mgx_filt_prev[franzosa_mgx_filt_prev < 0.0001] <- 0
prev = colSums(franzosa_mgx_filt_prev) 

franzosa_mgx_filt = franzosa_mgx[,prev >= round(nrow(franzosa_mgx)*10/100, digits=0)]

# take intersect also in IHMP data
intersecting.mgx = intersect(colnames(franzosa_mgx_filt), colnames(ihmp_mgx))
# all 355 intersect

print(paste0(length(intersecting.mgx), " taxa remain that are shared with IHMP"))

```

Excellent. 355 taxa is a reasonable number to perform this analysis on.

Now let's assess the distributions of taxa and log-transform if necessary, applying a single pseudocount to both datasets.

```{r distribution_check_mgx, echo=F, results=T, plot=T, warning=F, message=F }

# first, check each dataset's original distributions
franzosa_raw_mgx_hist = ggplot(reshape2::melt(as.matrix(franzosa_mgx_filt)))+
  ggridges::geom_density_ridges(aes(x=value, group=Var2, y=1),
                                  color="black", fill="white", alpha=0.2)+
  theme_minimal()+
  labs(x="Abundance", y="Frequency", title="Metagenomics\nFranzosa (Untransformed)")

ihmp_raw_mgx_hist = ggplot(reshape2::melt(as.matrix(ihmp_mgx)))+
  ggridges::geom_density_ridges(aes(x=value, group=Var2, y=1),
                                  color="black", fill="white", alpha=0.2)+
  theme_minimal()+
  labs(x="Abundance", y="Frequency", title="Metagenomics\nIHMP (Untransformed)")

# next, log-transform, but select a pseudocount that can be applied to both datasets
franzosa_mgx_pseudo = min(franzosa_mgx_filt[franzosa_mgx_filt!=0])/2
ihmp_mgx_pseudo = min(ihmp_mgx[ihmp_mgx!=0])/2
# select smaller of the 2
muller_mgx_pseudo = min(c(franzosa_mgx_pseudo, ihmp_mgx_pseudo))

franzosa_mgx_log = log(franzosa_mgx_filt+muller_mgx_pseudo, base=2)
ihmp_mgx_log = log(ihmp_mgx+muller_mgx_pseudo, base=2)


franzosa_log_mgx_hist = ggplot(reshape2::melt(as.matrix(franzosa_mgx_log)))+
  ggridges::geom_density_ridges(aes(x=value, group=Var2, y=1),
                                  color="black", fill="white", alpha=0.2)+
  theme_minimal()+
  labs(x="Log Abundance", y="Frequency", title="Metagenomics\nFranzosa (Log-transformed)")

ihmp_log_mgx_hist = ggplot(reshape2::melt(as.matrix(ihmp_mgx_log)))+
  ggridges::geom_density_ridges(aes(x=value, group=Var2, y=1),
                                  color="black", fill="white", alpha=0.2)+
  theme_minimal()+
  labs(x="Log Abundance", y="Frequency", title="Metagenomics\nIHMP (Log-transformed)")

# plot all
library("patchwork") %>% suppressMessages()
(franzosa_raw_mgx_hist | ihmp_raw_mgx_hist) / 
  (franzosa_log_mgx_hist | ihmp_log_mgx_hist)

```

Log-transforming improved the distributions of taxa in both datasets.

## Building models

Our objective is to predict the abundance of each metabolite using the taxa abundances.

I'm going to write a function that performs the following:

- Selects one metabolite (from the Franzosa dataset)

- Builds a model that predicts the metabolite's abundance using taxa abundances (using the Franzosa dataset). We're going to use random forest **OOB predictions**, which are as robust as multi-fold cross-validation, but **blazingly** faster.

- Builds a model that predicts scrambled (null) metabolite abundances

- Records the Spearman correlation of predictions from internal validation samples

- Applies the model to the IHMP dataset, and records the Spearman correlation coefficient

And repeat this process across all metabolites

```{r predictor_function, echo=F, results=T, plot=T, warning=F, message=F }

if(cache == F){
  
metabolite_predictor = function(train_mgx = franzosa_mgx_log,
         train_mbx = franzosa_mbx_log,
         test_mgx = ihmp_mgx_log,
         test_mbx = ihmp_mbx_log,
         metabolite = met,
         null = TRUE,
         model = algo,
         type = "oob", # only for ranger; otherwise 5xcv
         iter = iter){
  
  debug=F
  if(debug == T){
  train_mgx = franzosa_mgx_log
  train_mbx = franzosa_mbx_log
  test_mgx = ihmp_mgx_log
  test_mbx = ihmp_mbx_log
  null = TRUE
  metabolite = "X__propionate"
  model = "glmnet"
  type = "5xcv"
  iter = 1
  }
  
  # format data: make.names for feature names
  train_mgx = data.frame(train_mgx)
  test_mgx = data.frame(test_mgx)
  # scale metabolite
  train_mbx = scale(train_mbx)
  test_mbx = scale(test_mbx)
  
  # append metabolite of interest to training data
  train_mgx$metabolite = train_mbx[,metabolite]
    # build null data
  set.seed(iter)
  train_mgx_null = train_mgx %>%
    mutate(metabolite = sample(metabolite, replace=F))
  
  if(type != "oob"){
  # build real model
  set.seed(iter)
  trained_model <- caret::train(
    metabolite ~ .,
    data = data.frame(train_mgx),
    trControl = caret::trainControl(method="cv", 
                                    number = 5,
                                    savePred=T),
    metric = "RMSE",
    method = model,
    preProc = c("center", "scale")) %>% suppressWarnings()
  
  # build null model
  if(null == TRUE){
  set.seed(iter)
  null_model <- caret::train(
    metabolite ~ .,
    data = data.frame(train_mgx_null),
    trControl = caret::trainControl(method="cv", 
                                    number = 5,
                                    savePred=T),
    metric = "RMSE",
    method = model,
    preProc = c("center", "scale")) %>% suppressWarnings()
  }
  # make function to correlate final model predictions (with help from Grok)
extract_best_pred <- function(model) {
  if (!inherits(model, "train")) stop("Input must be a caret train object")
  best_tune <- model$bestTune
  tune_cols <- names(best_tune)
  best_preds <- model$pred
  for (col in tune_cols) {
    best_preds <- best_preds[best_preds[[col]] == best_tune[[col]], ]
  }
  # take average pred across folds
  best_preds = best_preds %>%
    group_by(rowIndex) %>%
    mutate(mean.pred = mean(pred)) %>%
    dplyr::select(mean.pred, obs, rowIndex) %>% distinct()
  return(best_preds)
}
# apply
best_pred_data_real <- extract_best_pred(trained_model)
  if(null == TRUE){
best_pred_data_null <- extract_best_pred(null_model)
}
      
  # extract Spearman cor
  internal_real_spearman = cor.test(best_pred_data_real$mean.pred,
                                    best_pred_data_real$obs, method="spearman")$estimate
    if(null == TRUE){
  internal_null_spearman = cor.test(best_pred_data_null$mean.pred,
                                    best_pred_data_null$obs, method="spearman")$estimate
    }
  
  # apply to real data
  predicted_real = predict(trained_model, data.frame(test_mgx))
    if(null == TRUE){
  predicted_null = predict(null_model, data.frame(test_mgx))
    }
  
  }
  # or build OOB ranger
  if(type == "oob"){

      set.seed(iter)
      trained_model <- ranger::ranger(metabolite ~., data.frame(train_mgx))
      set.seed(iter)
      null_model <- ranger::ranger(metabolite ~., data.frame(train_mgx_null))

      # extract Spearman cor
  internal_real_spearman = cor.test(trained_model$predictions,
                                    train_mgx$metabolite, method="spearman")$estimate
    if(null == TRUE){
  internal_null_spearman = cor.test(null_model$predictions,
                                    train_mgx_null$metabolite, method="spearman")$estimate
    }
  
  # apply to real data
  predicted_real = predict(trained_model, data.frame(test_mgx))$predictions
    if(null == TRUE){
  predicted_null = predict(null_model, data.frame(test_mgx))$predictions
    }
  }
  
  external_real_spearman = cor.test(predicted_real,
                                    test_mbx[,metabolite], method="spearman")$estimate
    if(null == TRUE){
  external_null_spearman = cor.test(predicted_null,
                                    test_mbx[,metabolite], method="spearman")$estimate
    }
  
  # collate into dataframe
    if(null == TRUE){
  data.frame(
    seed = iter,
    model = model,
    metabolite = metabolite,
    internal_real_spearman = internal_real_spearman,
    internal_null_spearman = internal_null_spearman,
    external_real_spearman = external_real_spearman,
    external_null_spearman = external_null_spearman)
    }else{
  data.frame(
    seed = iter,
    model = model,
    metabolite = metabolite,
    internal_real_spearman = internal_real_spearman,
    external_real_spearman = external_real_spearman)
    }
}

}
```

```{r ranger_loop,  echo=F, results=F, plot=T, warning=F, message=F }

metabolites = colnames(franzosa_mbx_log)
if(cache == F){
t1 <- Sys.time()
muller_ranger_output = do.call(rbind, lapply(1:15, function(iter){
      print(paste0(iter))
  do.call(rbind, parallel::mclapply(metabolites, function(met){
  metabolite_predictor(metabolite = met,
                       model="ranger",
                       type = "oob",
                     iter = iter)
  }))}))

t2 <- Sys.time()
t2-t1 # 37 min

saveRDS(muller_ranger_output, "../saved/muller_ranger_output.Rds")

}

muller_ranger_output = readRDS("../saved/muller_ranger_output.Rds")

```

```{r ranger_plot, echo=F, results=T, plot=T, warning=F, message=F }

# reorder metabolites
muller_ranger_output = muller_ranger_output %>%
  group_by(metabolite) %>%
  mutate(mean_internal_real = mean(internal_real_spearman),
         mean_internal_null = mean(internal_null_spearman),
         mean_external_real = mean(external_real_spearman),
         mean_external_null = mean(external_null_spearman)) 

muller_ranger_output_mean = muller_ranger_output %>%
  dplyr::select(metabolite,
                mean_internal_real, mean_internal_null, 
                mean_external_real,mean_external_null) %>% distinct()

print(paste0(subset(muller_ranger_output_mean, mean_external_real > 0.3) %>% nrow(), " metabolites with a Spearman rho > 0.3"))

ggplot(subset(muller_ranger_output_mean, mean_external_real > 0.3),
       aes(x=reorder(gsub("X__", "", metabolite), mean_external_real)))+
  # plot violins
  geom_violin(data = subset(muller_ranger_output, mean_external_real > 0.3), aes(y=internal_null_spearman), fill="grey", alpha=0.2)+
  geom_violin(data = subset(muller_ranger_output, mean_external_real > 0.3), aes(y=external_null_spearman), fill="grey", alpha=0.2)+
  geom_violin(data = subset(muller_ranger_output, mean_external_real > 0.3), aes(y=internal_real_spearman, fill="1"), color="black", alpha=0.2)+
  geom_violin(data = subset(muller_ranger_output, mean_external_real > 0.3), aes(y=external_real_spearman, fill="2"), fill=2, color="black", alpha=0.2)+
  # plot mean values
  geom_point(aes(y=mean_internal_null), shape=21, size=2, fill="grey")+
  geom_point(aes(y=mean_external_null), shape=21, size=2, fill="grey")+
  geom_point(aes(y=mean_internal_real, fill="1"), shape=21, size=2)+
  geom_point(aes(y=mean_external_real, fill="2"), shape=21, size=2)+ 
  theme_minimal()+theme(legend.position="none")+
  scale_y_continuous(breaks=seq(-6,7, by=1)/10)+
  coord_flip()+
  labs(y="Spearman ρ", x="")

```

Here, I've plotted the metabolites with a Spearman correlation coefficient above 0.3 using the external evaluation (IHMP) dataset. Results from the internal data are in red, while external data are in blue. Null performances (built and evaluated using internal and external data) are in grey. All models were built and evaluated 15 times.

The authors used a threshold of 0.3 as well, but added additional statistical validation for their definition of a metabolite that can be robustly predicted.

Note: the original study used only healthy microbiomes to build predictors, whereas I am focusing on IBD microbiomes. Despite this key difference, I've nonetheless validated that *urobilin* can be well predicted. Interestingly, this metabolite is also useful for discriminating healthy from IBD microbiomes, as I showed [here](https://pdobrano25.github.io/ml_website/hmp_validation.html#Feature_importances).


# **Extending** the analysis

Now, we'll ask whether other machine learning models are superior to random forest. The authors also evaluated lasso and elastic net. To save on computation time, I'm going to limit this benchmark to metabolites with external validation Spearman correlations > 0.3. We're also going to drop the null tests, since our goal is to evaluate whether these other models outperform random forest.

```{r model_loop,  echo=F, results=T, plot=T, warning=F, message=F }

subset.metabolites = subset(muller_ranger_output_mean, mean_external_real > 0.3)$metabolite

ml.models = c("gbm","glmnet","ranger","svmLinear","svmRadial","xgbTree")

if(cache == F){
  
t1 <- Sys.time()
muller_loop_output = do.call(rbind, lapply(1:15, function(iter){
  do.call(rbind, lapply(ml.models, function(algo){
        print(paste0(iter, " ", algo))
  do.call(rbind, parallel::mclapply(subset.metabolites, function(met){
  metabolite_predictor(metabolite = met,
                       model = algo,
                       type = "5xcv",
                       null = FALSE,
                       iter = iter)
  }))}))}))

t2 <- Sys.time()
t2-t1 # 9 hours total

saveRDS(muller_loop_output, "../saved/muller_loop_output.Rds")

}

muller_loop_output = readRDS("../saved/muller_loop_output.Rds")

```

```{r model_plots, echo=F, results=T, plot=T, warning=F, message=F }

# take mean of 15 iterations
muller_loop_output_mean = muller_loop_output %>% 
  group_by(model, metabolite) %>%
  mutate(mean_internal_real = mean(internal_real_spearman),
         mean_external_real = mean(external_real_spearman))

ggplot(muller_loop_output_mean[,c("model", "metabolite", "mean_internal_real")] %>% distinct(),
       aes(x=model, y=mean_internal_real))+
  geom_violin(draw_quantiles = 0.5)+
  ggbeeswarm::geom_beeswarm(shape=21, aes(fill=model))+
  theme_minimal()+theme(legend.position="none",
                        axis.text.x=element_text(angle=45,vjust=1, hjust=1))+
  labs(x="", y="Mean Spearman ρ", title="Internal Validation")|
ggplot(muller_loop_output_mean[,c("model", "metabolite", "mean_external_real")] %>% distinct(),
       aes(x=model, y=mean_external_real))+
  geom_violin(draw_quantiles = 0.5)+
  ggbeeswarm::geom_beeswarm(shape=21, aes(fill=model))+
  theme_minimal()+theme(legend.position="none",
                        axis.text.x=element_text(angle=45,vjust=1, hjust=1))+
  labs(x="", y="Mean Spearman ρ", title="External Validation")
```

It doesn't appear like anything is superior to random forest. Let's test this statistically using a mixed effects model.

```{r model_stats,  echo=T, results=T, plot=T, warning=F, message=F }

# set Random Forest (ranger) as the reference group
muller_loop_output$model = factor(muller_loop_output$model, levels=c("ranger", "gbm", "glmnet", "svmLinear", "svmRadial", "xgbTree"))

# build a mixed effects regression model
lmerTest::lmer(external_real_spearman ~ model + (1|metabolite), muller_loop_output) %>%
  summary() %>% coef()

```
The mixed effects model shows random forest is significantly superior (across all metabolites), except for svmRadial.

Now, we sort of biased these results by selecting the top metabolites (ρ > 0.3) identified by random forest. It's entirely possible that other models could have identified **different** metabolites better than random forest, which could have bumped up their overall performance. However, these 41 metabolites alone took 9 hours to compute.

## Feature importances

Since random forest was superior, let's go back to the first models we built using the ranger package. We'll extract out the feature importances for metabolites and correlate them with the taxa to identify their direction of association.

```{r ranger_feature_importances, echo=F, results=T, plot=T, warning=F, message=F }

# loop through metabolites; build predictor, extract importances, AND calculate spearman correlation

if(cache == F){
  
t1 <- Sys.time()

ranger.features = do.call(rbind, lapply(subset.metabolites, function(met){
    print(paste0(met))
  
    train_mgx = franzosa_mgx_log
    train_mbx = franzosa_mbx_log[,met]
  
    # build RF (take mean of 15 iter)
    train_mgx$metabolite = train_mbx
    
    rf.feature.loop = do.call(rbind, lapply(1:15, function(iter){
      
    set.seed(iter)
    trained_model <- ranger::ranger(metabolite ~., data.frame(train_mgx),
                                    importance = "impurity")
    rf.feature.importance = data.frame(importance = trained_model$variable.importance) %>%
      mutate(taxa = rownames(.),
             iter = iter,
             metabolite = met)
    }))
    # take mean
    rf.feature.loop.mean = rf.feature.loop %>%
      group_by(taxa) %>%
      mutate(importance = mean(importance)) %>%
      dplyr::select(taxa, importance) %>% distinct()
    
    # calculate correlations
    cors = reshape2::melt(cor(train_mgx)) %>%
      data.frame() %>%
      subset(Var1 == "metabolite") %>%
      mutate(taxa = Var2,
             cor = value)
    
    # merge
    merge(rf.feature.loop.mean, cors[,c("taxa", "cor")]) %>%
      mutate(metabolite = met)
    }))

  t2 <- Sys.time()
  t2-t1 # 2 min
  saveRDS(ranger.features, "../saved/ranger.features.Rds")
  
}

```

Let's plot a heatmap with features (taxa) and their associations with the top 41 metabolites.

We'll select only the top 20 taxa based on their mean importance value.

```{r feature_heatmap, echo=F, results=T, plot=T, warning=F, message=F }

ranger.features =  readRDS("../saved/ranger.features.Rds")

ranger.features.top.taxa = ranger.features %>%
  group_by(taxa) %>%
  mutate(mean.imp = mean(importance)) %>%
  dplyr::select(taxa, mean.imp) %>% distinct() %>%
  arrange(-mean.imp) %>%
  head(n=20)

reshape2::acast(subset(ranger.features, taxa %in% ranger.features.top.taxa$taxa),
                taxa ~ gsub("X__", "", metabolite), value.var="importance") %>%
  pheatmap::pheatmap()

```

There are a few metabolites with an feature importances worth noting.

First, **urobilin**, which I previously found to be associated with IBD microbiomes [here](https://pdobrano25.github.io/ml_website/hmp_validation.html#Feature_importances). Urobilin has a few features with outsized importance values towards the top right of the heatmap.

Second, **chenodeoxycholate** (CDCA), a primary bile acid, has a band of importance features. Primary bile acids, including CDCA, have been consistently found to be enriched in IBD, including in Franzosa's original [study](https://www.nature.com/articles/s41564-018-0306-4). 

Let's explore the features most strongly predictive of these two metabolites


```{r, feature_plot, echo=F, results=T, plot=T, warning=F, message=F }

ranger.features.top = subset(ranger.features,
                             metabolite %in% c("X__urobilin", "X__chenodeoxycholate"))


# plot
ggplot(ranger.features.top %>% arrange(-importance),
       aes(x=cor, y=importance))+
  geom_point(shape=21, aes(fill=cor, size=importance))+
  ggrepel::geom_label_repel(data = subset(ranger.features.top, importance > 200),
                           aes(x=cor, y=importance, label=taxa), size=3)+
  scale_fill_gradient2(high=scales::muted("red"),
                       mid="white",
                       low=scales::muted("blue"))+
  labs(x="Spearman ρ", y="Feature Importance")+
  theme_minimal()+theme(legend.position="none")+
  facet_wrap(~gsub("X__", "", metabolite), ncol=2) %>% suppressWarnings()

```

For **urobilin**, we see *Holdemania* is most strongly associated and predictive feature.

For **CDCA**, several taxa are negatively associated, including *Angelakisella*, *Pullichristensenella*, and an unclassified species annotated as *UBA866.* I'm unfamiliar with these taxa names.

[*Angelakisella*](https://www.ncbi.nlm.nih.gov/Taxonomy/Browser/wwwtax.cgi?id=1871018) is a genus in the Oscillospiraceae family. [*UBA866*](https://www.ncbi.nlm.nih.gov/Taxonomy/Browser/wwwtax.cgi?mode=Info&id=1952448&lvl=6&lin=f&keep=1&srchmode=5&unlock&filter=genome_filter) is annotated as a Ruminococcaceae. [*Pullichristensenella*](https://www.ncbi.nlm.nih.gov/Taxonomy/Browser/wwwtax.cgi?mode=Undef&id=3097446&lvl=3&keep=1&srchmode=1&unlock) is a candidatus (proposed, but not yet accepted) annotation for a genus in the Clostridia incertae sedis family.

# Conclusions

To conclude, we were able to predict *some* metabolite abundances using genus-level metagenomic data in a cohort of IBD patients. We employed a separate cohort to validate these predictions.

We found that random forest was superior (or at least non-inferior) to the other models.

We identified specific taxa that are predictive of certain metabolites, like urobilin and CDCA, which are relevant to IBD.
